---
title: "Project: Diagnostic Growth in UEC"
subtitle: "SU actions following steering group meeting on 03/09/2024"
date: 2024-09-17
format:
  html:
    html-table-processing: none
    # code-fold: true
    code-tools: true
    embed-resources: true
    grid:
      margin-width: 350px
  pdf: default
reference-location: margin
citation-location: margin
toc: true
toc-title: Contents
toc-location: left
theme: journal
fig-cap-location: top
number-sections: true
title-block-banner: '#f8be06'
title-block-banner-color: '#242322'
# mainfont: Gill Sans MT
# mainfont: Helvetica
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library("gt")
library("gtExtras")
library("here")
library("dplyr")
library("tidyr")
library("purrr")
library("readr")
library("ggplot2")
library("janitor")
library("stringr")
library("lubridate")

# FOR GRAPHIC FYEAR AXIS LABELS:
every_nth <-  function(n) {
  return(function(x) {x[c(TRUE, rep(FALSE, n - 1))]})
}

# 1. LOAD DATASETS --------------------------------------------

# FOR REGION LKP:
eric_2223 <- read_csv("https://files.digital.nhs.uk/FB/BE3AC8/ERIC%20-%20202223%20-%20Trust%20data.csv") |> 
  clean_names() |> 
  mutate(region = str_remove_all(commissioning_region, " COMMISSIONING REGION")) |> 
  select(trust_code, trust_name, region, trust_type)

list.files(here::here("data_raw"))
dq_multicoding <- readRDS(
  here("data_raw", "from_ncdr_trends_240917_dq_multicoding.rds")
  )
ts_provider_attds <- readRDS(
  here("data_raw", "from_ncdr_240917_ts_prov_att.rds")
  )
dq_provider_na_24_99 <- readRDS(
  here("data_raw", "from_ncdr_trends_240917_dq_prov_na_24_99.rds")
  )
dq_provider_represent <- readRDS(
  here("data_raw", "from_ncdr_240917_prov_represent.rds")
  ) 
# DEPENDS ON RESULTS OF THIS DOCUMENT:
df_preplot_trend <- readRDS(
  here("data_raw", "from_ncdr_trends_241003_df_preplot.rds")
  )

# 2. JOIN --------------------------------------------

dq_provider_represent <- dq_provider_represent |> 
  left_join(eric_2223, join_by(procode == trust_code))

# 3. PROVIDER SURVIVAL --------

ts_provider_survival <-  ts_provider_attds |>
  count(fyear) 

providers_surviving <- ts_provider_attds |>
  # MORE THAN AV OF 15 A DAY FOR A YEAR (EXCLUDES RVR ANOMALY)
  filter(n > 15*365) |>
  # NOW CLOSED (FROM MID 2023):
  filter(procode != "RVY") |>
  complete(fyear, procode) |>
  group_by(procode) |>
  filter(!is.na(sum(n))) |>
  ungroup() |>
  distinct(procode)

# 4. MULTICODING --------

dq_multicoding_sensitivity <- dq_multicoding |>
  cross_join(
    tibble(n_tests_of_type_in_record = c(2, 3, 4))
  ) |>
  cross_join(
    # MINIMUM PROPORTION OF CASES (AS PROPORTION OF 1-4) IN EACH YEAR:
    tibble(annual_instance_threshold = c(
      1 / 100000,
      1 / 50000,
      1 / 10000,
      1 / 5000,
      1 / 1000,
      1 / 500,
      1 / 100,
      1 / 50,
      1 / 10
    ))
  ) |>
  mutate(data = pmap(
    list(data, n_tests_of_type_in_record, annual_instance_threshold),
    \(df, y, z)
    df |>
      mutate(procode = str_sub(procode, 1, 3)) |> 
      filter(code_occurs == y) |>
      # WE WILL GET PRACTICE VARIATION ACROSS PROVIDERS, SO WHAT P DO WE PICK?
      # (% CASES IN ALL YEARS > z)
      filter(p > z) |>
      # NEXT FEW LINES ARE TO MAKE SURE ALL YEARS ACCOUNTED FOR:
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup()
  ))

dq_multicoding_provider_n <- dq_multicoding_sensitivity |>
  mutate(n_prov = map_vec(
    data,
    \(df)
    df |>
      distinct(procode) |>
      count() |>
      pull(n)
  )) |>
  mutate(n_prov_500 = map_vec(
    data,
    \(df)
    # MORE THAN 500 CASES EACH YEAR:
    df |>
      filter(n >= 500) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |>
  mutate(n_prov_50 = map_vec(
    data,
    \(df)
    # MORE THAN 50 CASES EACH YEAR:
    df |>
      filter(n >= 50) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |> 
  mutate(n_prov_20 = map_vec(
    data,
    \(df)
    # MORE THAN 20 CASES EACH YEAR:
    df |>
      filter(n >= 20) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |> 
  mutate(n_prov_2 = map_vec(
    data,
    \(df)
    # MORE THAN 1 CASE EACH YEAR:
    df |>
      filter(n >= 2) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |> 
  mutate(n_tests_of_type_in_record = as.character(n_tests_of_type_in_record)) |>
  mutate(invest_name = case_when(
    invest_code == "01" ~ "01 | X-ray",
    invest_code == "03" ~ "03 | Haematology",
    invest_code == "05" ~ "05 | Biochemistry",
    invest_code == "10" ~ "10 | Ultrasound",
    invest_code == "11" ~ "11 | MRI",
    invest_code == "12" ~ "12 | CT"
  ))

dq_multicoding_provider_lists_relaxed <- dq_multicoding_provider_n |> 
  filter(
    (invest_code == "01" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/1000)|
    (invest_code == "03" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/500)|
    (invest_code == "05" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/100)|
    (invest_code == "12" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/1000)
  ) |> 
  mutate(prov_list = map(
    data,
    \(df)
    df |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
  mutate(prov_list_500 = map(
    data,
    \(df)
    df |> 
      filter(n >= 500) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
  mutate(prov_list_50 = map(
    data,
    \(df)
    df |> 
      filter(n >= 50) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
   mutate(prov_list_20 = map(
    data,
    \(df)
    df |> 
      filter(n >= 20) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
   mutate(prov_list_4 = map(
    data,
    \(df)
    df |> 
      filter(n >= 5) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    ))

dq_providers_relaxed <-
  semi_join(
    providers_surviving,
  reduce(dq_multicoding_provider_lists_relaxed$prov_list_4, semi_join),
  join_by(procode)
  )

```

# Background

In Phase I of the project, we seek answers to five questions:

::: {.callout-note appearance="minimal"}
1.  How has the average number of tests per ED attendance changed over time?

2.  For each of the 20+ test types, how much more likely is it that a patient will receive a test in ED in year 2023/24 than in year 2019/20?

3.  What proportion of the growth in each test, can be attributed to: (a) increases in raw demand; (b) changes in case-mix; and (c) changes in the propensity to test (i.e. clinical practice)?

4.  How much (net) time does a test add to a patient’s stay in ED?

5.  How do tests influence decisions to admit?
:::

To address these questions, we will rely on SUS A&E and EC data sets. Whilst these are relatively high-quality sources, we are aware of substantial variation in recording practices, across the population of provider trusts and over time. If overlooked, this variation is likely to affect the *reliability* and *validity* of our results.

It is therefore our intention to use data sampling (and use different samples to answer the different types of question). We will include a provider in a sample if the quality of their data submissions meet our requirements for the given question.

We expect this approach to improve the reliability and validity of our results. However, since we also wish make general statements about the effect of diagnostic testing on urgent and emergency care, we must ensure that our sample and our population (in each case) have similar properties. Our sample of providers must be *representative* of the population of providers.

There is a trade-off to be made here. The highest quality data, taken from a small number of providers, may increase the reliability of our results. But the smaller the sample of providers, the less likely it is to be representative of the population of providers.

Mindful of this trade-off, we set out our recommendations for the sample of providers to be used in each analysis.

# Recommendations: Q1 {#sec-rec-1}

::: {.callout-note appearance="minimal"}
1.  *How has the average number of tests per ED attendance changed over time?*
:::

This analysis covers a period of 12 years and therefore spans A&E and EC data sets (each of which has a slightly different structure).

We identified `r nrow(providers_surviving)` providers with provider codes that survived across the 12 year study period and that, in addition, passed basic data quality checks. These `r nrow(providers_surviving)` providers represent the total population of providers that we might have used for our analysis had data quality been universally high.

Providers were included in our sample if their data submissions met the quality required for this question. Here, the study period is relatively long but we have few variables to examine. We are looking to see that:

1.  Throughout the study period, providers have had consistently high levels of recording of the investigation fields.

2.  For attendances in which multiple investigations of the same type have been carried out, providers have recorded each instance of the investigation (and have been consistent in this practice over the whole study period).

We established that `r nrow(dq_providers_relaxed)` providers met these criteria. The providers we intend to use for our sample are given in @tbl-sample-q1. In the adjacent tab, @tbl-representation suggests that this sample of `r nrow(dq_providers_relaxed)` providers is widely representative of the population of providers. We note that the sample includes higher proportions of larger providers and providers in London, and lower proportions of smaller providers and providers in the East of England.

::: panel-tabset
## Sample of Providers (for Q1)

```{r}
#| label: tbl-sample-q1
#| tbl-cap: "Sample of Providers Recommended for Question 1"
#| echo: false
#| message: false

dq_providers_relaxed |> 
  left_join(eric_2223, join_by(procode == trust_code)) |> select(1,2) |> 
  rename(Provider_Code = procode, Provider_Name = trust_name) |>
    arrange(Provider_Code) |> 
  mutate(id = row_number(), .before = Provider_Code) |> 
  mutate(Provider_Name = str_to_title(Provider_Name)) |> 
  mutate(Provider_Name = str_replace_all(Provider_Name, "Nhs", "NHS")) |> 
  gt() |> 
    fmt_auto() |> 
   tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(c(Provider_Name, Provider_Code))
  ) |> 
  tab_options(
    data_row.padding = px(1),
      table.font.size = "small"
    )
  
```

## Sample Representation (for Q1)

```{r}
#| label: tbl-representation
#| tbl-cap: "Is the Sample Representative of the Wider Provider Population?"
#| echo: false
#| message: false

table_representation <- tibble(
  data = list(
    dq_provider_represent |>
      semi_join(providers_surviving, join_by(procode)),
    dq_provider_represent |>
      semi_join(dq_providers_relaxed, join_by(procode))
  )
) |>
  mutate(n_providers = map_dbl(data, \(df)
  df |>
    count(procode) |>
    nrow())) |>
  # THIS WILL BE SIZE IN 2023/24:
  # (FOR OTHER VARS WE'LL LOOK AT MEAN ACROSS 5 YR ECDS TIME PERIOD)
  mutate(size = map(data, \(df)
  df |>
    filter(fyear == "2023/24") |> 
    count(fyear, procode, wt = n) |>
    group_by(procode) |>
    # MEAN OF A PROVIDER'S SIZE OVER 5 YEARS:
    reframe(mean_n = mean(n)) |>
    mutate(size_group = case_when(
      mean_n < 70e3 ~ "att < 70k",
      mean_n >= 70e3 & mean_n < 95e3 ~ "att 70-94k",
      mean_n >= 95e3 & mean_n < 130e3 ~ "att 95-129k",
      mean_n >= 130e3 ~ "att > 130k",
      T ~ NA_character_
    )) |>
    mutate(size_group = factor(
      size_group,
      levels = c("att < 70k", "att 70-94k", "att 95-129k", "att > 130k")
    )) |>
    count(size_group) |>
    mutate(p = n / sum(n)) |>
    select(-n) |>
    pivot_wider(names_from = size_group, values_from = p)
  
  )) |>

  mutate(arrive_ambulance = map_dbl(data, \(df)
  df |>
    count(fyear, is_ambulance, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_ambulance == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(admitted = map_dbl(data, \(df)
  df |>
    count(fyear, is_adm, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_adm == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(age_under_18 = map_dbl(data, \(df)
  df |>
    count(fyear, age_under_18, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(age_under_18 == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(age_75_plus = map_dbl(data, \(df)
  df |>
    count(fyear, age_75_plus, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(age_75_plus == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(urban = map_dbl(data, \(df)
  df |>
    count(fyear, is_urban, wt = n) |>
    # BECAUSE LOTS OF NAS IN 2023/24:
    filter(fyear != "2023/24") |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_urban == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(deprived_20pc = map_dbl(data, \(df)
  df |>
    count(fyear, imd_quint_1, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(imd_quint_1 == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(region = map(data, \(df)
  df |>
    count(fyear, region, wt = n, sort = T) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    group_by(region) |>
    summarise(av = round(mean(p), 3)) |>
    ungroup() |>
    pivot_wider(names_from = region, values_from = av) |>
    clean_names() |>
    rename_with(~ paste0("region_", .x, recycle0 = TRUE)) |>
    identity())) |>
  unnest(region) |>
  unnest(size) |>
  select(-data) |>
  mutate(zz = if_else(n_providers > 50, "population", "sample"), .before = n_providers) |>
  pivot_longer(cols = !starts_with("zz"), names_to = "variable") |>
  pivot_wider(names_from = zz, values_from = value)

table_representation |> 
  select(variable, population, sample) |> 
  mutate(across(2:3, ~if_else(is.na(.), 0, .))) |>
  rowwise() |> 
  mutate(diff = sample - population) |> 
  gt() |> 
  opt_row_striping(row_striping = F) |> 
  fmt_auto() |> 
  tab_style(
    style = list(
      cell_text(transform = "capitalize", weight = "bold")
    ),
    # different location
    locations = cells_column_labels(everything())
  ) |> 
  fmt_percent(rows = c(2:18), decimals = 1) |>
  tab_row_group(
    label = "Patient demographics:",
    rows =  str_detect(variable, "age|urban|depriv|region_")
  ) |> 
  tab_row_group(
    label = "Case-mix-related:",
    rows =  str_detect(variable, "arrive_|admitted")
  ) |> 
  tab_row_group(
    label = "Provider Size (attendances 23/24):",
    rows =  str_detect(variable, "att")
  ) |> 
  tab_row_group(
    label = "",
    rows =  str_detect(variable, "n_prov")
  ) |>
  tab_style(
  style = cell_text(weight = "bold"),
  locations = cells_row_groups(groups = c(1, 2, 3, 4))
) |> 
    sub_values(values = -71, replacement = "") |> 
   gt_highlight_cols(sample, fill = "grey", alpha = 0.1) |> 
  tab_options(
    data_row.padding = px(1),
      table.font.size = "small"
    )

```
:::

# Provisional results for Q1

@fig-trend shows provisional results for Question 1, based on data from the sample of `r nrow(dq_providers_relaxed)` providers identified above. It appears that, overall, test rates have doubled over the study period, from an average of 2 tests per attendance in 2012/13 to 4 tests per attendance in 2023/24. The rate of increase for admitted patients was uneven and appears to have slowed over the period (now normalising after the pandemic-related spike 2020/2021). For the 7 in 10 patients who are are not admitted after an ED attendance, the increase was fairly consistent to 2020. Again, the rate of increase seems to have slowed following the spike in test rates (seen across the majority of test types) in 2020/21.

```{r}
#| label: fig-trend
#| echo: false
#| fig-cap: "Estimated average number of tests per ED attendance, in England, between 2012/13 and 2023/24. Source: SUS A&E and EC datasets."
#| message: false


df_preplot_trend |>
  ggplot(aes(fyear, rate)) +
  geom_line(aes(col = disdest, group = disdest)) +
  geom_point(aes(col = disdest, group = disdest)) +
  geom_blank(aes(y = 0)) +
  scale_color_manual(values = c("indianred3", "dodgerblue", "purple4")) +
  labs(x = "Financial year")+
  theme_bw()+
  labs(y = "Average number of recorded tests per ED attendance")+
  theme(
    legend.position = "none",
    axis.ticks = element_blank()
    )+
  facet_wrap(vars(disdest))+ # , scales = "free_y"
  scale_x_discrete(breaks = every_nth(n=4))+
  scale_y_continuous(breaks = seq(0,9, by = 2))


# preplot_trend_by_invest <- readRDS("from_ncdr_trends_241003_preplot_by_invst.rds")
# 
# options(scipen = 999)
# 
# preplot_trend_by_invest |>
#   ggplot(aes(fyear, rate)) +
#   geom_line(aes(col = disdest, group = disdest)) +
#   geom_point(aes(col = disdest, group = disdest)) +
#   geom_blank(aes(y = 0)) +
#   labs(x = "Financial year") +
#   theme_bw() +
#   labs(subtitle = "Vertical axis: Average number of investigations per A&E attendance") +
#   theme(
#     axis.title.y = element_blank()
#   ) +
#   facet_wrap(vars(invst_type), scales = "free_y") + # , scales = "free_y"
#   scale_x_discrete(breaks = every_nth(n = 3)) +
#   # scale_y_continuous(breaks = seq(0,9, by = 2))+
#   NULL
```

# Recommendations: Qs 2-5

```{r}
#| include: false

dqual_ecds_only <- readRDS(
  here("data_raw", "from_ncdr_ecds_only_240917_dq_isna.RDS")
  )
dqual_ecds_only_provider_key <- readRDS(
  here("data_raw", "from_ncdr_ecds_only_240917_dq_provider.RDS")
  ) |> 
   # NOT PROVIDERS WITH ARRIVAL MODE = NULL > 5% (SEE EDA ON NCDR DS SERVER):
  filter(!procode %in% c(
    "RCX",
    "RQM",
    "RJ7",
    "RAX",
    "RQX",
    "RVY",
    "RP5",
    "RRV"
  ))

providers_surviving_ecds_only <- ts_provider_attds |>
    # NOT PROVIDERS WITH ARRIVAL MODE = NULL > 5% (SEE EDA ON NCDR DS SERVER):
  filter(!procode %in% c(
    "RCX",
    "RQM",
    "RJ7",
    "RAX",
    "RQX",
    "RVY",
    "RP5",
    "RRV"
  )) |> 
  # MORE THAN AV OF 15 A DAY FOR A YEAR (EXCLUDES RVR ANOMALY):
  filter(n > 15*365) |>
  # NOW CLOSED (FROM MID 2023):
  filter(procode != "RVY") |>
  filter(fyear %in% c("2019/20", "2023/24")) |>
  complete(fyear, procode) |>
  group_by(procode) |>
  filter(!is.na(sum(n))) |>
  ungroup() |>
  distinct(procode)


providers_key_90 <- dqual_ecds_only_provider_key |> 
# TO KEEP CONSISTENT WITH ABOVE CHANGE TO % NA:
  mutate(across(starts_with("x"), ~ 1 - .)) |> 
  filter(!if_any(starts_with("x"), ~ is.na(.))) |> 
  rowwise() |> 
  mutate(xmean = mean(c_across(x2019_20: x2023_24))) |> 
  ungroup() |> 
  filter(if_all(starts_with("x"), ~ . < .10)) 

providers_key_85 <- dqual_ecds_only_provider_key |> 
# TO KEEP CONSISTENT WITH ABOVE CHANGE TO % NA:
  mutate(across(starts_with("x"), ~ 1 - .)) |> 
  filter(!if_any(starts_with("x"), ~ is.na(.))) |> 
  rowwise() |> 
  mutate(xmean = mean(c_across(x2019_20: x2023_24))) |> 
  ungroup() |> 
  filter(if_all(starts_with("x"), ~ . < .15)) |> 
  anti_join(
    providers_key_90, join_by(procode)
  ) 


provider_sample <- bind_rows(
  providers_key_90,
  # providers_key_85
) |> 
  select(procode)

dq_provider_represent_ecds_only <- dq_provider_represent |>
  filter(fyear %in% c("2019/20", "2023/24")) |> 
  filter(na_arrival == 0) |> 
  filter(covid_exclusion == 0) |> 
  select(-c(na_arrival, covid_exclusion))
```

::: {.callout-note appearance="minimal"}
2.  *For each of the 20+ test types, how much more likely is it that a patient will receive a test in ED in year 2023/24 than in year 2019/20?*
3.  *What proportion of the growth in each test, can be attributed to: (a) increases in raw demand; (b) changes in case-mix; and (c) changes in the propensity to test (i.e. clinical practice)?*
4.  *How much (net) time does a test add to a patient’s stay in ED?*
5.  *How do tests influence decisions to admit?*
:::

To carry out these analyses, we require only the EC data set, and only two years (provisionally, 2019/20 and 2023/24). We identified `r nrow(providers_surviving_ecds_only)` providers with provider codes that survived across this study period and that, in addition, passed basic data quality checks. These `r nrow(providers_surviving_ecds_only)` providers represent the total population of providers that we might have used for our analysis had data quality been universally high.

Providers were included in our sample if they had all three of *Der_EC_Investigation_All, chiefcompgrp,* and *acuity* complete for 90% of records. (Relaxing this requirement to 85% gave us only four more providers and no improvement in representation.) For Question 4, we also required that time-related variables were largely complete.

We established that `r nrow(providers_key_90)` providers met these criteria. The providers we intend to use for our sample are given in @tbl-sample-q2. In the adjacent tab, @tbl-representation-ecds suggests that this sample of `r nrow(providers_key_90)` providers is generally representative of the population of providers. We note that the sample includes higher proportions of larger providers and providers in the Midlands. One of the seven regions, the East of England, is not represented.

::: panel-tabset
## Sample of Providers (for Q2-5)

```{r}
#| label: tbl-sample-q2
#| tbl-cap: "Sample of Providers Recommended for Questions 2-5"
#| echo: false
#| message: false

provider_sample |> 
  left_join(eric_2223, join_by(procode == trust_code)) |> select(1,2) |> 
  rename(Provider_Code = procode, Provider_Name = trust_name) |> 
  arrange(Provider_Code) |> 
   mutate(id = row_number(), .before = Provider_Code) |> 
  mutate(Provider_Name = str_to_title(Provider_Name)) |> 
  mutate(Provider_Name = str_replace_all(Provider_Name, "Nhs", "NHS")) |> 
  gt() |> 
    fmt_auto() |> 
   tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(c(Provider_Name, Provider_Code))
  ) |> 
  tab_options(
    data_row.padding = px(1),
      table.font.size = "small"
    )
```

## Sample Representation (for Q2-5)

```{r}
#| label: tbl-representation-ecds
#| tbl-cap: "Is the Sample Representative of the Wider Provider Population?"
#| echo: false
#| message: false

table_representation_ecds_only <- tibble(
  data = list(
    dq_provider_represent_ecds_only |>
        semi_join(providers_surviving_ecds_only, join_by(procode)),
    dq_provider_represent_ecds_only |>
        semi_join(provider_sample, join_by(procode))
  )
) |>
  mutate(n_providers = map_dbl(data, \(df)
  df |>
    count(procode) |>
    nrow()
  )) |>
   mutate(f_year = map(data, \(df)
  df |>
    count(fyear, wt = n) |>
    mutate(p = n / sum(n)) |>
    filter(fyear == "2023/24") |> 
  select(-n) |>
  pivot_wider(names_from = fyear, values_from = p)
    )) |>
  mutate(size = map(data, \(df)
  df |>
    filter(fyear == "2023/24") |>
    count(fyear, procode, wt = n) |>
    group_by(procode) |>
    # MEAN OF A PROVIDER'S SIZE OVER 1 YEARS:
    reframe(mean_n = mean(n)) |>
    mutate(size_group = case_when(
      mean_n < 70e3 ~ "att < 70k",
      mean_n >= 70e3 & mean_n < 95e3 ~ "att 70-94k",
      mean_n >= 95e3 & mean_n < 130e3 ~ "att 95-129k",
      mean_n >= 130e3 ~ "att > 130k",
      T ~ NA_character_
    )) |>
    mutate(size_group = factor(
      size_group,
      levels = c("att < 70k", "att 70-94k", "att 95-129k", "att > 130k")
    )) |>
    count(size_group) |>
    mutate(p = n / sum(n)) |>
    select(-n) |>
    pivot_wider(names_from = size_group, values_from = p)
  
  )) |>
  mutate(arrive_ambulance = map_dbl(data, \(df)
  df |>
    count(fyear, is_ambulance, wt = n) |>
    # filter(fyear != "2023/24") |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_ambulance == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(admitted = map_dbl(data, \(df)
  df |>
    count(fyear, is_adm, wt = n) |>
    # filter(fyear == "2023/24") |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_adm == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(age_under_18 = map_dbl(data, \(df)
  df |>
    count(fyear, age_under_18, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(age_under_18 == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(age_75_plus = map_dbl(data, \(df)
  df |>
    count(fyear, age_75_plus, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(age_75_plus == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(urban = map_dbl(data, \(df)
  df |>
    count(fyear, is_urban, wt = n) |>
    # BECAUSE LOTS OF NAS IN 2023/24:
    filter(fyear != "2023/24") |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_urban == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(deprived_20pc = map_dbl(data, \(df)
  df |>
    count(fyear, imd_quint_1, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(imd_quint_1 == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(region = map(data, \(df)
  df |>
    count(fyear, region, wt = n, sort = T) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    group_by(region) |>
    summarise(av = round(mean(p), 3)) |>
    ungroup() |>
    pivot_wider(names_from = region, values_from = av) |>
    clean_names() |>
    rename_with(~ paste0("region_", .x, recycle0 = TRUE)) |>
    identity())) |>
  unnest(f_year) |>
  # mutate(type = map(data, \(df)
  #                   df |>
  #                     count(fyear, trust_type, wt = n, sort = T) |>
  #                     group_by(fyear) |>
  #                     mutate(p = n/sum(n)) |>
  #                     group_by(trust_type) |>
  #                     summarise(av = round(mean(p), 3)) |>
  #                     ungroup() |>
  #                     pivot_wider(names_from = trust_type, values_from = av) |>
  #                     clean_names() |>
  #                     rename_with(~ paste0("type_", .x, recycle0 = TRUE)) |>
  #                     identity()
  # )) |>
  # unnest(type) |>
  unnest(region) |>
  unnest(size) |>
  # unnest(type) |>
  select(-data) |>
  mutate(zz = if_else(n_providers > 50, "population", "sample"), .before = n_providers) |>
  pivot_longer(cols = !starts_with("zz"), names_to = "variable") |>
  pivot_wider(names_from = zz, values_from = value)

table_representation_ecds_only |> 
  select(variable, population, sample) |> 
  mutate(across(2:3, ~if_else(is.na(.), 0, .))) |> 
  rowwise() |> 
  mutate(diff = sample - population) |> 
  gt() |> 
  opt_row_striping(row_striping = F) |> 
  fmt_auto() |> 
  tab_style(
    style = list(
      cell_text(transform = "capitalize", weight = "bold")
    ),
    # different location
    locations = cells_column_labels(everything())
  ) |> 
  fmt_percent(rows = c(2:19), decimals = 1) |>
  tab_row_group(
    label = "Patient demographics:",
    rows =  str_detect(variable, "age|urban|depriv|region_")
  ) |> 
  tab_row_group(
    label = "Case-mix-related:",
    rows =  str_detect(variable, "arrive_|admitted")
  ) |> 
  tab_row_group(
    label = "Provider Size (attendances 23/24):",
    rows =  str_detect(variable, "att")
  ) |> 
  # tab_row_group(
  #   group = "Provider-related",
  #   rows =  str_detect(variable, "type_|min_|max_|av_")
  # ) |> 
  tab_row_group(
    label = "",
    rows =  str_detect(variable, "n_prov|2023")
  ) |>
  tab_style(
  style = cell_text(weight = "bold"),
  locations = cells_row_groups(groups = c(1, 2, 3, 4))
) |> 
    sub_values(values = -86, replacement = "") |> 
   gt_highlight_cols(sample, fill = "grey", alpha = 0.1) |> 
  tab_options(
    data_row.padding = px(1),
       table.font.size = "small"
    )
```
:::
