---
title: "Growth in Diagnostic Testing and Its Impact on Emergency Departments"
#"Growth in Diagnostic Testing<br>and Its Impact on<br>Emergency Departments"
subtitle: "<br>Steven Wyatt and Andrew Jones | The Strategy Unit | November 2024"
# author: 
#   - name: Steven Wyatt
#     email: swyatt@nhs.net
#     corresponding: TRUE
#     affiliation: The Strategy Unit
#   - name: Andrew Jones
#     email: swyatt@nhs.net
#     corresponding: FALSE
# date: 2024-11-29
format:
  html:
    html-table-processing: none
    # code-fold: true
    code-tools: true
    embed-resources: true
    grid:
      margin-width: 280px
  pdf: default
reference-location: margin
citation-location: margin
toc: true
toc-title: Contents
toc-location: left
theme: journal
fig-cap-location: top
number-sections: false
title-block-banner: '#f8be06'
title-block-banner-color: '#242322'
# mainfont: Gill Sans MT
# mainfont: Helvetica
editor_options: 
  chunk_output_type: console
---

# Executive Summary

Andrew Jackson's suggested/preferred/desired template:

Last time we said this, now we're saying this. (Where is the new work consistent with past work - where different.)

\[He is intending to use this as further evidence to suggests that 4HR target is not appropriate\]

# Introduction

Over the last decade, growth in diagnostic testing in emergency departments (EDs) has been rapid. Earlier analysis conducted by the Strategy Unit (2019, 2023) suggests that this growth may have implications for the timeliness of patient care, for decisions on whether to admit a patient, and for crowding in EDs.

::: column-margin
For the sake of readability, we use the term, “test”, to refer to a unit of diagnostic activity. We acknowledge that some of these “tests” might also be therapeutic procedures.
:::

In this report, we revisit.... / update selected analyses from these past reports , in order to...

we seek answers to SIX questions:

::: {.callout-note appearance="minimal" icon="false"}
## Questions Addressed in This Report:

<br>

1.  How has the number of tests per ED attendance changed over time?

2.  Are patients more likely to receive tests in EDs today than they were in 2019/20?

3.  What is driving test growth in EDs?

    <!--(Is it changes to demand, case-mix, or practice?)  -->

4.  How much time do tests add to a patient’s stay in an ED?

5.  How does time in ED change according to the number of tests conducted?

6.  How do tests in EDs influence decisions to admit?
:::

It is hoped that answers to these questions .... \[how do we wish to frame this?\]

A second phase of this project, for which findings are due in the spring of 2025, will disregard the boundary between EDs and inpatient beds. And look at the impact of diagnostic growth on urgent and emergency care a inpatient beds. Rather than looking at the im... might measure the total time that patients spend in contact with hospital services, starting when they enter the UEC pathway.

# Analytical Approach {#sec-approach}

The analyses we present here investigates the effect of diagnostic testing on major emergency (Type 1) departments. To address the questions above, we relied on SUS A&E and EC data sets. These are patient-level data sets...

### The use of provider sampling

Whilst SUS A&E and EC data sets are relatively high-quality sources, we observed considerable variation in recording practices, both across the population of provider trusts and over time. If overlooked, this variation would adversely affect both the *reliability* and *validity* of our results. \[links to meanings?\]

Therefore, to ensure that data met our quality requirements for a given question, we analyses we carry out in on samples of providers These samples were non-random: we included a provider in a sample if the quality of their data submissions met our requirements for the given question.

Whilst there were several different types of question we aimed to answer, we identified two samples. one sample of providers for Question 1 and another sample for Questions 2 to 5.

As This sampling approach should improve the reliability and validity of our results. However, we also wish make general statements about the effect of diagnostic testing on urgent and emergency care. We might only use results from our sample to make general comments if our sample of providers is *representative* of the population of providers. That is to say, our sample and our population (in each case) should have similar characteristics.

With a representative sample, we can be more confident that our findings for the sample are also applicable to the population.

There is a trade-off to be made here. The larger the sample of providers we select, the more likely it will be that this

The more providers we capture in our sample, the more likely the sample is to be representative of the population of providers.

A large sample of providers is more likely to be representative of the population of providers. But the larger we make our sample, the more we must compromise on data quality the lower we data quality threshold becomes, which is in turn likely to decrease the the reliability of our results.

Details of the sampling approaches, and measures of representation, and our on the trade-off above, can be found in Appendix 1.

### Attendances examined {#sec-approach-attd}

For providers within our samples, we included all unplanned attendances of persons with a postcode in England. We excluded cases were a person left the ED before being treated, and persons streamed to other care services. In addition, we excluded cases where the patient died during an attendance and those where the patient was found to be dead on arrival at the ED.

### Question-specific approaches

Details of our approach to each distinct question can be found in the expandable "Methods and Limitations" boxes, in the corresponding subsections of the Analysis section, below.

# Analysis

```{r}
#| label: setup
#| include: false

library("gt")
library("here")
library("dplyr")
library("tidyr")
library("forcats")
# library("purrr")
# library("readr")
library("ggplot2")
library("janitor")
library("stringr")
library("gtExtras")
library("lubridate")
```

### Perhaps some context - if time?

e.g. Describe the pop attending EDs show the casemix in both years, age, condtion... can't exclude all casemix effects

## Q1. How has the number of tests per ED attendance changed over time?

We start by looking at how the average number of tests per ED attendance changed over time. The analysis here covers a period of 12 years. Other summary details...?

::: {.callout-note collapse="true" appearance="minimal" icon="false"}
# Q1: Methods and Limitations {style="size: 20"}

###### <br>Methods

As this analysis covers a 12-year period, we had to employ both A&E and EC data sets.

We identified 28 providers with provider codes that survived across the 12 year study period and that passed our data quality criteria. See appendix...

The numerator for our "test rate" in a given year was the total number of tests performed across our provider sample. Where a patient received multiple tests of the same type, we counted these as multiple tests. The denominator for our "test rate" in a given year was the total number of ED attendances across our provider sample that matched the criteria given in @sec-approach-attd.

###### Limitations

Sample of 28. Providers who Multiple testing may not be consistent across providers or may multiple represent an error.

A&E and EC data sets (each of which has a slightly different structure).

Incompatibility of data sets -Did what we could to ensure consistency
:::

@fig-trend shows the change in test rates for the sample of 28 providers identified above. It appears that, overall (as shown in the left-hand panel), test rates doubled over the study period, from an average of 2 tests per attendance in 2012/13 to 4 tests per attendance in 2023/24.

For the 3 in 10 patients who are are admitted after an ED attendance, the recorded rate of increase was uneven but appears to have slowed over the period. During the pandemic period, ED attendances generally fell, while evidence suggests that persons who did attend ED were sicker. We suggest that the trend is now normalising after the short-term spike in test rates above the trend.

For the 7 in 10 patients who are are not admitted after an ED attendance, the increase was fairly consistent to 2020. Again, the rate of increase seems to have slowed following the spike in test rates (seen across the majority of test types) in 2020/21.

<br>

::: panel-tabset
## Graphical Presentation

```{r}
#| label: fig-trend
#| echo: false
#| fig-cap: "Estimated average number of tests per ED attendance, in England, between 2012/13 and 2023/24. Source: SUS A&E and EC datasets."
#| message: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

df_preplot_trend <- readRDS(
  here("data_raw", "from_ncdr_trends_241003_df_preplot.rds")
  ) |> 
  mutate(disdest = fct_relevel(disdest, "overall (all destinations)")) |> 
  mutate(disdest = fct_relevel(disdest, "non-admitted patients", after = 1))

# FOR GRAPHIC FYEAR AXIS LABELS:
every_nth <-  function(n) {
  return(function(x) {x[c(TRUE, rep(FALSE, n - 1))]})
}

df_preplot_trend |>
  ggplot(aes(fyear, rate)) +
  geom_line(aes(col = disdest, group = disdest)) +
  geom_point(aes(col = disdest, group = disdest)) +
  geom_blank(aes(y = 0)) +
  scale_color_manual(values = c("indianred3", "dodgerblue", "purple4")) +
  labs(x = "Financial year")+
  theme_bw()+
  labs(y = "Average number of recorded tests per ED attendance")+
  theme(
    legend.position = "none",
    axis.ticks = element_blank()
    )+
  facet_wrap(vars(disdest))+ # , scales = "free_y"
  scale_x_discrete(breaks = every_nth(n=4))+
  scale_y_continuous(breaks = seq(0,9, by = 2))
```

## Tabular Presentation

If we want numerators/denoms, we'll need to reproduce the source object for the graph, which means going back to NCDR
:::

## Q2. Are patients more likely to receive tests today than they were in 2019/20?

How much more likely was a patient to receive a test in an ED in 2023/24 compared to 2019/20?

Start off with some intro text from Diagnostics report 2023

Are patients more likely to receive tests today than they were in 2019/20?

## Q3. What is driving test growth in EDs?

Q3. Is the growth in EDs being driven by changes to demand, case-mix, or practice?

## Q4. How much time do tests add to a patient's stay in the ED?

How much (net) time does a test add to a patient’s stay in ED?

## Q5. How do tests influence decisions to admit?

More text here.

# Discussion

# Appendix 1

Sampling and SQL queries ? Collapsable?

<!-- To carry out these analyses, we require only the EC data set, and only two years (provisionally, 2019/20 and 2023/24). We intend to use different samples to answer the different types of question). We will include a provider in a sample if the quality of their data submissions meet our requirements for the given question. We expect this approach to  -->

### Sampling for Q1 (from trends dq)

Given that we will be covering a relatively long period, spanning two data sets (with slightly different structures), it will be important to understand the quality and consistency of recording practices, both over time and by provider. In earlier [work in this area](https://www.strategyunitwm.nhs.uk/publications/waiting-times-and-attendance-durations-english-ae-departments), we based our analysis on subsets of providers who had a history of high quality data. For this analysis, it is likely that we will, again, have to work with data from a subset providers.

There is more - will have to look...

We identified XXX providers with provider codes that survived across the 12 year study period and that, in addition, passed basic data quality checks. These XXX providers represent the total population of providers that we might have used for our analysis had data quality been universally high.

Providers were included in our sample if their data submissions met the quality required for this question. Here, the study period is relatively long but we have few variables to examine. We are looking to see that:

1.  Throughout the study period, providers have had consistently high levels of recording of the investigation fields.

2.  For attendances in which multiple investigations of the same type have been carried out, providers have recorded each instance of the investigation (and have been consistent in this practice over the whole study period).

We established that XX providers met these criteria. The providers we intend to use for our sample are given in @tbl-sample-q1. In the adjacent tab, @tbl-representation suggests that this sample of XXX providers is widely representative of the population of providers. We note that the sample includes higher proportions of larger providers and providers in London, and lower proportions of smaller providers and providers in the East of England.

## Sampling for Q2 (from ECDS DATA qual qmd)

One solution to this problem may be to use other variables to understand case-mix (and may even be more valid measures). For example, the combination of *acuity* (composed of 5 levels) and Chief Complaint Group (15 groups) should allow us to account for differences in case-mix. Both variables have \<13% NULL values in both years when looking at the data set as a whole. The completion rates would likely improve further if we were to examine a subset of providers.

It is possible, then, that the quality of a provider’s submission of these two variables, along with *Der_EC_Investigation_All*, may determine whether we include the provider in our analyses. We will address this idea next.

we deduce that 16 providers can claim to have **all 3 fields** **complete** for \>90% of attendances, in both years. If we were to relax the threshold to \>85%, we get just four more providers.

### How representative is the provider sample?

We identified 102 providers with provider codes that were present in the two years of interest and that, in addition, passed basic data quality checks. These providers will represent the total population of providers that we might have used for our analyses. Given the data quality issues, we have selected a sample of providers to improve the reliability and validity of our results. However, since we wish to make general statements about about the effect of diagnostic testing on urgent and emergency care, we must ensure that our samples are representative of the population of providers.

More deprived, and older population, with more arrivals by ambulance results in more admissions. We will control for all of these factors - what difference does this make…?
