---
title: "Growth in Diagnostic Testing and Its Impact on Emergency Departments"
#"Growth in Diagnostic Testing<br>and Its Impact on<br>Emergency Departments"
subtitle: "<br>Steven Wyatt and Andrew Jones | The Strategy Unit | November 2024"
# author: 
#   - name: Steven Wyatt
#     email: swyatt@nhs.net
#     corresponding: TRUE
#     affiliation: The Strategy Unit
#   - name: Andrew Jones
#     email: swyatt@nhs.net
#     corresponding: FALSE
# date: 2024-11-29
format:
  html:
    html-table-processing: none
    # code-fold: true
    code-tools: true
    embed-resources: true
    grid:
      margin-width: 280px
  pdf: default
reference-location: margin
citation-location: margin
toc: true
toc-title: Contents
toc-location: left
theme: journal
fig-cap-location: top
number-sections: false
title-block-banner: '#f8be06'
title-block-banner-color: '#242322'
# mainfont: Gill Sans MT
# mainfont: Helvetica
editor_options: 
  chunk_output_type: console
---

# Executive Summary

# Introduction

[**ADD POLICY COMMENT - NOTE TRADE OFF. RECOMENDED INTENTIONAL GROWTH, NOT JUST RISING TO MEET DIAGNOSTIC CAPACITY?**]{.underline}

Over the last decade, growth in diagnostic testing in emergency departments (EDs) has been rapid. Earlier analysis conducted by the Strategy Unit ([2019](https://www.strategyunitwm.nhs.uk/publications/waiting-times-and-attendance-durations-english-ae-departments), [2023](https://www.strategyunitwm.nhs.uk/sites/default/files/2023-05/MDSN%20Diagnostic_growth_report%20Apr2023.pdf)) suggests that this growth may have implications for the timeliness of patient care, for decisions on whether to admit a patient, and for crowding in EDs.

::: column-margin
For the sake of readability, we use the term, “test”, to refer to a unit of diagnostic activity. We acknowledge that some of these “tests” might also be therapeutic procedures.
:::

In this report, we update and expand our analysis to assess the extent to which the trends identified in earlier reports have continued. We explore growth in diagnostic testing in EDs in England, and consider the implications of this growth for the timeliness and disposition of attendances. We sought to answer six questions:

::: {.callout-note appearance="minimal" icon="false"}
#### Questions on diagnostic testing in EDs that we address in this report:

<br>

1.  How has the number of tests per attendance changed over time?

2.  Can the rise in test rates be explained by changes to casemix?

3.  What is driving the growth in testing?

4.  How much time do tests add to a patient’s attendance?

5.  How does time in ED change according to the number of tests conducted?

6.  How do tests influence decisions to admit?
:::

We start by describing our analytical approach, before addressing each question in turn. Detailed information about the methods and limitations are provided for each analysis.

<!--In the second phase of this project, for which findings are due in the spring of 2025, we will disregard the boundary between EDs and inpatient beds. And look at the impact of diagnostic growth on urgent and emergency care a inpatient beds. Rather than looking at the im... might ///Increased diagnostic testing means that patients spend less time in contact with hospital services, even ifpatients an increase in time in ED.///we measure the total time that patients spend in contact with hospital services, starting when they enter the UEC pathway.-->

# Analytical Approach {#sec-approach}

[**C- AJ**]{.underline}

The analyses we present here investigate the consequences of increased diagnostic testing within major emergency (Type 1) departments. To address the questions above, we relied on SUS A&E and EC data sets. Whilst these are high-quality, patient-level sources, we observed considerable variation in recording practices, both across the population of provider trusts and over time. If overlooked, this variation would adversely affect both the [reliability](https://en.wikipedia.org/wiki/Reliability_(statistics)) and [validity](https://en.wikipedia.org/wiki/Validity_(statistics)#Reliability) of our results.

#### The use of provider sampling {#sec-approach-sampling}

To increase the chances that our findings would be reliable and valid for a given question, we conducted each analysis on data from a sample of providers. These provider samples were non-random: we included a provider in a sample if the quality of their data submissions met our requirements for the given question.

Whilst we addressed six distinct questions, we were able to identify two different samples. We identified a sample of 28 providers with which we could address Question 1. A second sample, of 16 providers, could be used for the analysis of Questions 2 to 6.

Whilst sampling should improve the reliability and validity of our results, we cannot make general statements about the effect of testing on EDs *unless* our sample of providers is *representative* of the population of providers. That is to say, the providers in our sample(s) and population should have similar characteristics.

<!--With a representative sample, we can be more confident that our findings for the sample are also applicable to the population.-->

There is a trade-off to be made here. The larger the sample of providers we select, the more likely it will be that this

The more providers we capture in our sample, the more likely the sample is to be representative of the population of providers.

The larger our sample of providers, the more likely it is to be representative of the population of providers. But the larger we make our sample, the more we must compromise on data quality, which will likely to decrease the the reliability of our results.

Details of the sampling methods, our approach to the trade-off, and the measures of representation can be found in @sec-appendix.

#### Excluded attendances {#sec-approach-attd}

For providers within our samples, the analyses cover all unplanned attendances of persons with an English postcode, with some exceptions. We excluded cases were a person left the ED before being treated, and persons streamed to other care services. In addition, we excluded cases where the patient died during an attendance and those where the patient was found to be dead on arrival at the ED.

#### Question-specific approaches

Details of our approach to each distinct question can be found in the expandable "Methods and Limitations" boxes, in the corresponding subsections of the Analysis section, below.

#### Reproducibility

The computer code used to perform this analysis can be found a [GitHub repository](https://github.com/The-Strategy-Unit/diagnostic_growth_uec).

# Analysis

```{r}
#| label: setup
#| include: false

library("gt")
library("here")
library("dplyr")
library("tidyr")
library("scales")
library("forcats")
library("ggrepel")
# library("purrr")
# library("readr")
library("ggplot2")
library("janitor")
library("stringr")
library("gtExtras")
library("lubridate")
```

#### Perhaps some context - if time?

**D - AJ**

e.g. Describe the pop attending EDs show the casemix in both years, age, condtion... can't exclude all casemix effects

## A1. How has the number of tests per ED attendance changed over time?

::: {#nte-q1-meth .callout-note collapse="true" appearance="minimal" icon="false"}
# A1: Method and Limitations (click to unhide)

<br>**Method**

To be more precise, we estimated how the *average* number of tests per ED attendance has changed over time. As this analysis covers a period of 12 years - starting with the financial year 2012/13 and concluding with 2023/24 - we had to employ both A&E and EC data sets.

We identified 28 providers with provider codes that survived across the 12 year study period and that passed our data quality criteria. See @sec-appendix.

The numerator for our "test rate" in a given year was the total number of tests performed across our provider sample. Where a patient received multiple tests of the same type, we counted these as multiple tests. The denominator for our "test rate" in a given year was the total number of ED attendances across our provider sample that matched the criteria given in @sec-approach-attd.

**Limitations**

Our aim is to make general statements about the effect of diagnostic testing on urgent and emergency care. Due to concerns over data quality, we employed sampling in this analysis. Our results come from 28 of 99 providers whose provider codes remained unchanged over the 12-year period. As the 28 providers were chosen on the basis of their data quality, this constitutes a non-random sample of providers. Study conclusions drawn from non-random samples are at higher risk of bias. As noted in @sec-approach-sampling, we attempted to mitigate the risk of bias by confirming that the sample was generally representative of the population of providers in various dimensions. It is, however, possible that we unknowingly omitted important dimensions for this particular question, and that our findings are less generalisable than we suggest.

There are two additional points worth noting. Firstly, in order to complete this analysis, we were obliged to unite A&E and EC data sets, each of which has a slightly different structure. We used the A&E data set for financial years between 2012/13 and 2018/19 and the EC data set between 2019/20 and 2023/24. Whilst we attempted to ensure consistency across data sets, it may be that there are some differences

The second important point is that we included only those providers who appeared to record multiple instances of distinct investigations (and who appeared to have been consistent in this practice over the whole study period). It is, however, possible that these providers were internally inconsistent in recording practices over the period or that, in some cases, the recording of multiple tests of the same type represented an error. \[Could reference our data analysis here\]
:::

**A1: Results**

@fig-trend shows the change in test rates for the sample of 28 providers identified above. It appears that, overall (as shown in the left-hand panel), test rates doubled over the study period, from an average of 2.1 tests per attendance in 2012/13 to 4.2 tests per attendance in 2023/24. This equates to a growth rate of 6.5% per annum. However, the growth in tests per attendance has reduced in recent years, from 9.5% per annum between 2012/13 to 2018/19, to 3.1% per annum between 2018/19 and 2023/24.

During the pandemic period, ED attendances fell. Evidence suggests that persons who did attend EDs were sicker, leading to more tests per patient. We suggest that the trend is now normalising after the short-term spike in test rates above the trend.

Although, over the study period, testing rates for non-admitted patients were lower than for patients who were admitted, the *increase* in testing rates was greater for non-admitted patients (8.1% vs 4.6% per annum, respectively).

<br>

::: panel-tabset
#### Graphical Presentation

```{r}
#| label: fig-trend
#| echo: false
#| fig-cap: "Estimated average number of tests per ED attendance, in England, between 2012/13 and 2023/24. Source: SUS A&E and EC data sets."
#| column: page
#| fig-width: 10
#| fig-asp: 0.618
#| fig-align: center
#| out-width: "85%"


# #| fig-align: center
# #| column: page
# #| fig-width: 8
# #| fig-asp: 0.618
# #| out-width: "90%"

df_preplot_trend <- readRDS(
  here("data_raw", "from_ncdr_trends_241003_df_preplot.rds")
  ) |> 
  mutate(disdest = fct_relevel(disdest, "overall (all destinations)")) |> 
  mutate(disdest = fct_relevel(disdest, "non-admitted patients", after = 1))

# FOR GRAPHIC FYEAR AXIS LABELS:
every_nth <-  function(n) {
  return(function(x) {x[c(TRUE, rep(FALSE, n - 1))]})
}

df_preplot_trend |>
  ggplot(aes(fyear, rate)) +
  geom_line(aes(col = disdest, group = disdest)) +
  geom_point(aes(col = disdest, group = disdest)) +
  geom_blank(aes(y = 0)) +
  scale_color_manual(values = c("indianred3", "dodgerblue", "purple4")) +
  labs(x = "Financial year")+
  theme_bw()+
  labs(y = "Average number of recorded tests per ED attendance")+
  theme(
    legend.position = "none",
    axis.ticks = element_blank()
    )+
  facet_wrap(vars(disdest))+ # , scales = "free_y"
  scale_x_discrete(breaks = every_nth(n=4))+
  scale_y_continuous(breaks = seq(0,9, by = 2))
```

#### Tabular Presentation

If we want numerators/denoms, we'll need to reproduce the source object for the graph, which means going back to NCDR
:::

## A2. Can the rise in test rates be explained by changes to casemix?

The number and types of test that a patient receives in ED will be influenced by the patient’s characteristics and presenting conditions. The term "casemix" can be used to describe the combination of these factors. In this section of analysis, we explore whether patients are more or less likely to receive a test in 2023/24 than in 2019/20, having taken account of the patient casemix. The results are expressed as casemix-adjusted odds ratios, which is the odds of a patient receiving a test in 2023/24 divided by the odds in 2019/20, having adjusted for changes in casemix between the two years. We perform these calculations for 32 tests. These 32 tests account for over 99% of all tests recorded in EDs.

::: {#nte-q2-meth .callout-note collapse="true" appearance="minimal" icon="false"}
# A2: Method and Limitations (click to unhide)

<br>**Method**

We used the SUS EC data set and identified 16 providers with provider codes that survived across the study period and that passed our data quality criteria for this question. For further details of our sampling approach and data quality criteria see @sec-approach-sampling and @sec-appendix.

We built 32 distinct models to predict casemix-adjusted odds ratios (as specified above) for 32 (of 49) test types that appear in the EC data set. These 32 tests account for over 99% of all tests recorded in EDs.

In terms of data preparation, we removed March from both 2019/20 and 2023/24, to avoid any bias to our results that may have arisen due to the onset of the pandemic in March 2020.

Approximately 4% of the 3 million records had missing values in a key field (our "key fields" are given in @sec-appendix). Our specified model type will not accept missing values, so we chose to delete the records with missing values in our key fields.

We performed logistic regressions using generalised additive models (GAMs) with mixed-effects. All 32 models were specified in the same way, with the only difference being the binary dependent variable (whether the given test was carried out). Our fixed effects included the variable of interest, financial year, and several time-related variables to control for season, weekday, and time of arrival. We included variables to adjust for casemix (as far as possible): *age* (as a spline-based smooth) interacting with *sex*; a re-engineered *chief complaint* variable; *acuity*; *arrival mode*; and *referral source*.

(It is important to note that the *chief complaint* and *acuity* variables are recorded as part of an initial assessement ([NHSE, 2022](https://www.england.nhs.uk/guidance-for-emergency-departments-initial-assessment/#:~:text=chief%20complaint%20or%20suspected%20diagnosis,initially%20assessed%20by%20a%20clinician)). They are akin to presentation codes. Whilst the diagnosis codes may give greater detail about a patient's condition, diagnosis is often a causal consequence of testing and it would therefore be inappropriate to use diagnosis-related information to predict whether a patient receives a given test.)

We treated the provider variable as a random effect. This allowed us to control for factors that vary, at random, across providers. Moreover, the values produced by this random effect can be thought of as sample values from a larger population of values. As the providers in our data set were, in fact, a sample of all providers, this approach suits our case well.

We ran each model on the same sample of 1 million attendance records.

**Limitations**

The findings of this analysis are based on a sample of 16 providers (of 102 providers whose provider codes remained unchanged over the study period). These findings are therefore subject to the same sampling-related limitations we identified for Q1 (see @nte-q1-meth).

We have labelled our estimates as "casemix-adjusted" as we controlled for a range of variables that might indicate a patient's need for a test. However, the clinician who assesses a patient will have far more information on that patient's condition than we have access to.

the fact th the fact that a clinician requested (or did not request) a test is likely to be indicative of a patient's condition and because we don't Because we don't have access to all the information used by the clinician, we can only partially control for casemix.

As we are missing, the fact that a clinician is likely to be indicative of a patient's condition. There may be many more variables that influence whether a test is given than we have access to. and the fact that a clinician requested (or did not request) a test, itself, is also likely to be indicative of a patient's condition. Because we are unable to control for unrecorded variables, we can only partially control for casemix, and it's possible that more (or less) of the change we've seen is due to casemix.

when deciding whether or not to order a test for a patients, clinician

We did not take into account the fact that a clinician requested (or did not request) a test. Because we are unable to control for these unrecorded variables, we cannot fully adjust our estimates for casemix, and our estimates may be biased. it's possible that more (or less) of the change we've seen is due to casemix.

and whilst it is true to say that

It is true to say this may only be a partial adjustment for casemix.

Clinicians have information, when deciding whether to order a test for a patients. Because we are unable to control for these unrecorded variables, we cannot fully adjust our estimates for casemix, and our estimates may be biased. it's possible that more (or less) of the change we've seen is due to casemix.

This information is also likely to be indicative of a patient’s condition. There is some that cannot be controlled for.

Finally, as noted in the Methods section, above, we decided to delete approximately 4% of records from our sample that had missing values for any of the variables used in the model. We expect this decision to have a limited impact on our results. However, as it reduced our sample size, the standard errors around our model coefficients have marginally increased as a consequence. There is also the potential for some bias in our results if these missing data were not missing completely at random.
:::

**A2: Results**

@fig-odds-sans-sero suggests that for most tests provided in EDs, patients were more likely to receive the test in 2023/24 than they were in 2019/20. In many cases, the increased odds of receiving a test were substantial. The adjusted odds of receiving the following tests increased by between 50% and 100%: hepatic function panel; CT scan; bone profile; analysis of arterial blood gases and pH; D-dimer assay; creatine kinase measurement; and MRI scan. Moreover, the adjusted odds of receiving c-reactive protein; urine sent for culture situation; serum lactate measurement; wound microscopy culture and sensitivities; lipid panel; and serologic tests more than doubled. 

For 6 of the 32 tests, the odds of receiving the test *fell* between 2019/20 and 2023/24. These tests were: erythrocyte sedimentation rate measurement, group and save, electrocardiography, serum lipase measurement, urinalysis, and human chorionic gonadotropin measurement.

We can compare the increased odds of testing estimated here, with similar analysis covering the period 2012/13 to 2019/20. This comparison suggests that the increase in the casemix-adjusted odds of a patient receiving a CT scan has fallen, from +19% per annum between 2012/13 and 2019/20, to +16% per annum between 2019/20 and 2023/24. There has been a similar slowdown in the rate of increase for MRI scans (from 25% per annum between 2012/13 and 2019/20, to 17% per annum between 2019/20 and 2023/24).

[**CHECK NUMBER OF ATTENDANCES WITH TEST - AND NOT JUST TESTS**]{.underline}

::: panel-tabset
## Graphic<br>(Excl. serological testing)

```{r}
#| label: fig-odds-sans-sero
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 10
#| fig-asp: 0.618
#| out-width: "85%"
 
# #| fig-width: 8
# #| fig-asp: 0.618
# #| out-width: "90%"

df_odds_results <- bind_rows(
  readRDS(here("data", "from_ncdr_growth_attrb_v3_1to16.rds")),
  readRDS(here("data", "from_ncdr_growth_attrb_v3_17to32.rds"))
) |>
  # THE NUMBER OF 2023/24 ATTENDANCES IN NESTED DFS FOR ODDS MODEL:
  mutate(att_2023_24 = 0.566e6) |>
  mutate(p = tests_2023_24 / att_2023_24)


# df_odds_results |>
  # filter(InvestigationDescription != "Serologic test") |> 
  # ggplot() +
  # geom_hline(yintercept = 1, lty = "dashed", colour = "grey20") +
  ## geom_hline(yintercept = 2, lty = "dashed", colour = "grey20") +
  # geom_col(aes(reorder(InvestigationDescription, odds), odds), alpha = .8) +
  # theme_bw() +
  # theme_minimal() +
  # coord_flip() +
  # scale_y_continuous(
  #   breaks = seq(0, 4, by = 1),
  #   labels = c(
  #     "",
  #     # "0.5x\nas likely in\n2023/24",
  #     "Equally\nas likely in\n2023/24",
  #     # "1.5x\nas likely in\n2023/24",
  #     "2x\nas likely in\n2023/24",
  #     "3x\nas likely in\n2023/24",
  #     "4x\nas likely in\n2023/24"
  #     ),
  #   position = "right"
  # ) +
  # theme(
  #   axis.title.y = element_blank(),
  #   plot.margin = margin(5, 20, 5, 5),
  #   panel.grid.minor = element_blank()
  # ) +
  # labs(y = "Case-mix-adjusted odds ratio of test (2023/24 vs 2019/20)\n")


df_odds_results |>
    filter(InvestigationDescription != "Serologic test") |>
  ggplot(aes(odds, p)) +
  geom_vline(xintercept = 1, lty = "dashed", colour = "grey20") +
  # theme_bw() +
  theme_minimal() +
  geom_point() +
  geom_blank(aes(x = 0, y = 0)) +
  geom_text_repel(
    aes(label = str_wrap(InvestigationDescription, 20)),
    size = 2,
  ) +
  scale_y_continuous(
    limits = c(0, 0.5),
    labels = scales::percent_format()
  ) +
  scale_x_continuous(
    breaks = seq(0, 4, by = 1),
    labels = c(
      "",
      # "0.5x\nas likely in\n2023/24",
      "Equally as likely\nin 2023/24",
      # "1.5x\nas likely in\n2023/24",
      "2x as likely\nin 2023/24",
      "3x as likely\nin 2023/24",
      "4x as likely\nin 2023/24"
      ),
    position = "top"
  ) +
  labs(
    x = "Case-mix-adjusted odds ratio of test (2023/24 vs 2019/20)\n",
    y = "Proportion of attendances involving test (2023/24)\n"
  )+
  theme(
        plot.margin = margin(5, 20, 5, 5)
  )+
  NULL

```

## Graphic<br>(Incl. serological testing)

```{r}
#| label: fig-odds
#| echo: false
#| message: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

df_odds_results |>
  # filter(InvestigationDescription != "Serologic test") |> 
  ggplot() +
  geom_hline(yintercept = 1, lty = "dashed", colour = "grey20") +
  # geom_hline(yintercept = 2, lty = "dashed", colour = "grey20") +
  geom_col(aes(reorder(InvestigationDescription, odds), odds), alpha = .8) +
  theme_bw() +
  theme_minimal() +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 5),
    labels = c(
      "",
      # "0.5x\nas likely in\n2023/24",
      # "Equally\nas likely in\n2023/24",
      # "1.5x\nas likely in\n2023/24",
      "5x\nas likely in\n2023/24",
      "10x\nas likely in\n2023/24",
      "15x\nas likely in\n2023/24",
      "20x\nas likely in\n2023/24"
      ),
    position = "right"
  ) +
  theme(
    axis.title.y = element_blank(),
    plot.margin = margin(5, 20, 5, 5),
    panel.grid.minor = element_blank()
  ) +
  labs(y = "Case-mix-adjusted odds ratio of test (2023/24 vs 2019/20)\n")
```

## Table

```{r}
#| label: tab-odds
#| echo: false
##| fig-cap: "Having adjusted for casemix, how much more likely it is that a patient will receive a test in an ED in 2023/24 compared to 2019/20? Source: ECDS."
#| message: false
##| fig-align: center
#| column: page
##| fig-width: 8
##| fig-asp: 0.618
##| out-width: "90%"

df_odds_results |>
  select(InvestigationDescription, att_2023_24, tests_2023_24, odds) |>
  mutate(tests_2023_24 = as.integer(tests_2023_24)) |> 
  gt() |>
  cols_label(
    InvestigationDescription = "Test name",
    att_2023_24 = "Attendances in sample<br>2023/24",
    tests_2023_24 = "Tests in sample<br>2023/24",
    odds = "Odds ratio of test (casemix adj.)<br>2023/24 vs. 2019/20", 
        .fn = md
  ) |> 
  fmt_auto() |>
  tab_options(
    data_row.padding = px(1),
    table.font.size = "small"
  ) 

```
:::

## A3. What is driving test growth in EDs?

In this section, we seek to decompose the growth in ED tests between 2019/20 to 2023/24 into to three main effects:

::: column-margin
Strictly speaking, we examined growth in ED attendances with at least one test, rather than growth in ED tests. The potential effect of this is discussed in the A3 Limitations section.
:::

-    Changes in demand: the number of patients presenting at ED

-    Changes in casemix: the acuity and complexity of patient presentations

-   Changes in practice: the ways in which ED respond to patients who present

Because these changes occur simultaneously, they interact in complex ways to influence the number of tests delivered. We separate out this interaction or compounding effect as a separate quantity. This, often modest, residual effect can be thought of as the difference between the total observed change, and the sum of the three individual effects. 

We present results for 32 tests. As noted, these 32 tests account for over 99% of all tests recorded in EDs.

::: {.callout-note collapse="true" appearance="minimal" icon="false"}
# A3: Method and Limitations (click to unhide)

<br>**Method**

Once again, this analysis relies on the ECDS and the 16 providers with provider codes that both survived across the study period and passed our data quality criteria for this question. For further details of our sampling approach and data quality criteria see @sec-approach-sampling and @sec-appendix.

For each of the 32 tests covered in A2, we set out to find the percentage growth in attendances with one or more of the specified test ($T$), as a function of the growth due to: changes in raw demand ($D$), changes in casemix ($C$), and changes driven by all other factors ($P$). (We expect $P$ to be dominated by the influence of clinical practice, but it may also contain other factors, such as the availability of resources.)

We formulated a multiplicative model

$1+T = (1+D)(1+C)(1+P)$ . \[1\]

Expanding the right hand side and solving for $T$ we get:

$T = D + C + P + [DC + DP + CP + DCP]$ . \[2\]

Thus, we are able to isolate the main effects for growth due to changes in demand, casemix, and practice. The terms within the square brackets are interaction effects (providing, for example, the effect of casemix for different levels of demand). In most cases, the influence of these interaction effects is likely to be negligible.

We have defined $T$ as,

$T = t_{23/24}/t_{19/20} -1$ , \[3\]

where $t$ is the number of attendances with one or more of the specified test in the financial year.

We might also define $D$, independently, as,

$D = a_{23/24}/a_{19/20} -1$ , \[4\]

where $a$ is the total number of attendances (both with and without a test) in the financial year.

We can then move to a calculation of $P$ which will rely on predictions from our logistic regression models from A2.

We will imagine a counterfactual scenario in which there was no change in casemix between 2019/20 and 2023/24 (i.e. we saw identical sets of patients attend in both years). In this case, both $D$ and $C$ will be equal to zero, and Equation 2 becomes:

$T_{cf} = P$ . \[5\]

where,

$T_{cf} = t'_{23/34(cf)} / t'_{19/20} -1$ . \[6\]

We can find $t'_{23/34(cf)}$ and $t'_{19/20}$ , and therefore $T_{cf}$ and thus $P$, using predictions from the logistic regression models from A2. (For our counterfactual scenario, we make predictions having re-labelled the 2019/20 attendances as 2023/24 attendances.)

We can then use Equation 2 and solve for C:

$C = (T - D - P - DP)/(1 + D + P + DP)$. \[7\]

Finally, we can find the interaction effects in Equation 2. We will collect them under the term, $X$, giving us:

$T = D + C + P + X$ , \[8\]

and

$X = T-D-C-P$ . \[9\]

The main graphical output of this analysis reflects Equation 8, but for the fact we chose to collect $P$ and $X$ terms in one panel. We have labelled this panel "growth due to changes in practice", as the influence of the $X$ term is often negligible.

**Limitations**

The findings of this analysis are based on a sample of 16 providers (of 102 providers whose provider codes remained unchanged over the study period). These findings are therefore subject to the same sampling-related limitations we identified for Q1 (see @nte-q1-meth).

Moreover, the analysis relies on the logistic regression models we built for Q2, hence the limitations we acknowledged in that analysis (see @nte-q2-meth) also apply here.

Demand growth is based on sample.

Unobserved casemix -as per A2.

It is growth in ED attendances with at least one test, and not growth in ED tests. For tests that are, in all but rare circumstances, only provided once per attendance, this approach will ... for others there will be some discrepancy.
:::

**A3: Results**

Of the total growth in the number of tests between 2019/20 and 2023/24, XX% can be attributed to changes in demand, YY% to changes in casemix, and ZZ% to changes in practice, with the remainder attributable to interactions between these terms.

Whilst the provider sample used in this analysis is representative of all providers in England in many ways, the providers in this subset experienced higher levels of demand growth than most between 2019/20 and 2023/24. Therefore, growth in tests due to demand increases is likely to be higher in these trusts than in England as a whole. However, the balance of growth attributable to casemix and practice changes *is* likely to be generalisable. On this basis, the analysis would suggest that the vast majority of the growth in tests per attendance is attributable to changes in practice rather than changes in casemix. 

In light of earlier analysis, which covered the period from 2010/11 to 2015/16, our findings here suggest that the driver of test growth in ED has changed in recent years. In the earlier analysis, we found that approximately one third of the growth in tests per attendance was attributable to changes in casemix. ED casemix became more complex and acute between 2010/11 and 2015/16, but it does not appear to have changed substantially between 2019/20 and 2023/24. This might explain, in part, the slowdown in the growth in tests per attendance seen since 2019/20.

Of course, some tests are more common than others. When we consider both test frequency and the change in the casemix-adjusted odds of a test, we can identify those tests where changes in practice have had a material impact on the total number of tests carried out. These are C-reactive protein tests, hepatic function panel tests, and CT scans. 

As medical practice and technology develops, and tests become more readily available, clinicians are able to replace older tests with more modern and (likely) more accurate tests. This might explain, for example, why the casemix-adjusted odds of CT and MRI scans has increased whilst there has been minimal change for plain radiography. And this same substitution effect may explain the different trends in electrocardiography and troponin measurement. Changes in coding practices may also play a part.

```{r}
#| include: false

df_preplot_growth_components <- df_odds_results |>
  mutate(d0 = 1, .before = dT) |>
  relocate(dC, .before = dP) |>
  pivot_longer(cols = matches("^d|^abs"), names_to = "component", values_to = "value") |>
  select(-tests_2023_24) |>
  mutate(value_abs = tests_2019_20 * value) |>
  mutate(value_order = if_else(component == "d0", value_abs, NA_real_)) |>
  mutate(value = if_else(component == "d0", NA_real_, value)) |>
  mutate(component = case_when(
    component == "d0" ~ "Attds with test\n19/20",
    component == "dT" ~ "Growth in attds with test\n19/20 to 23/24",
    component == "dD" ~ "Growth driven by demand",
    component == "dC" ~ "Growth driven by case-mix",
    component == "dP" ~ "Growth driven by practice",
    component == "dI" ~ "Growth due interactions",
    T ~ NA_character_
  )) |>
  mutate(component = forcats::as_factor(component)) |>
  mutate(value_abs = as.integer(value_abs)) |>
  group_by(InvestigationDescription) |>
  mutate(value_order = max(value_order, na.rm = T)) |>
  ungroup() |>
  mutate(value_abs = round(value_abs / 1000)) |>
  mutate(value_label_abs = if_else(is.na(value), str_c(value_abs, "k"), NA_character_)) |>
  mutate(value_label_p = if_else(value >= 0, str_c("+", round(value * 100), "%"), str_c(round(value * 100), "%")))


```

::: panel-tabset
## Percentage differences

```{r}
#| label: fig-attrib-perc
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-height: 10
#| out-width: "90%"

# #| fig-asp: 0.618

df_preplot_growth_components |>
  ggplot(aes(reorder(InvestigationDescription, value_order), value_abs)) +
  geom_col(
    aes(
      reorder(InvestigationDescription, value_order), value_abs,
      fill = component
    ),
    position = "identity",
    alpha = 0.4
  ) +
  geom_text(
    aes(
      label = value_label_abs,
      vjust = ifelse(value_abs >= 0, -0.4, 1.4),
      angle = 270
    ),
    size = 1.5,
    colour = "grey20"
  ) +
  geom_text(
    aes(
      label = value_label_p,
      vjust = ifelse(value_abs >= 0, -0.4, 1.4),
      angle = 270
    ),
    size = 1.5,
    colour = "grey20"
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, lty = "dashed") +
  theme_minimal() +
  theme(
    axis.ticks = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 5),
    panel.grid.minor = element_blank(),
    # panel.spacing = unit(1, "lines")
  ) +
  scale_y_continuous(
    limits = c(-50, 250)
  ) +
  facet_wrap(vars(component), ncol = 6) +
  labs(y = "\nED attendances (thousands) with 1+ test of specified type",
       caption = "*Demand growth reflects the demand growth in our sample")
```

## Absolute differences (for sample)

```{r}
#| label: fig-attrib-abs
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-height: 10
#| out-width: "90%"

# #| fig-asp: 0.618

df_preplot_growth_components |>
  mutate(value_label_abs = str_c(value_abs, "k")) |>
  ggplot(aes(reorder(InvestigationDescription, value_order), value_abs)) +
  geom_col(
    aes(reorder(InvestigationDescription, value_order), value_abs, fill = component),
    position = "identity",
    alpha = 0.4
  ) +
  geom_text(
    aes(
      label = value_label_abs,
      vjust = ifelse(value_abs >= 0, -0.4, 1.4),
      angle = 270
    ),
    size = 1.8,
    colour = "grey20"
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, lty = "dashed") +
  theme_minimal() +
  theme(
    axis.ticks = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 5),
    panel.grid.minor = element_blank(),
    # panel.spacing = unit(1, "lines")
  ) +
  scale_y_continuous(
    limits = c(-50, 250)
  ) +
  facet_wrap(vars(component), ncol = 6) +
  labs(y = "\nED attendances (thousands) with 1+ test of specified type")


```

## Table

To come.
:::

## A4. How much time do tests add to a patient's stay in the ED?

[**J - SW (intro text and description of graphic )**]{.underline}

How much (net) time does a test add to a patient’s stay in ED?

::: {.callout-note collapse="true" appearance="minimal" icon="false"}
## A4: Method and Limitations

<br>**Method**

This analysis again uses data from the 16 providers that passed our consistency and data quality criteria for A2-A6. (For further details of our sampling approach and data quality criteria see @sec-approach-sampling and @sec-appendix.) We use only records from the 2023/24 financial year here.

We removed records with times that appeared to be poorly recorded. These included attendances in which the arrival time was noted as occuring after the assessment time, and those in which the total time in the ED was recorded to be less than zero minutes. We also limited this analysis to attendances where the patient is recorded to have stayed in the ED for less than 30 hours.

Approximately 6.7% of the 1.8 million records had missing values in a key field (our "key fields" are given in @sec-appendix). Our specified model type will not accept missing values, so we chose to delete the records with missing values in our key fields.

Overall, with these exclusions, we removed 6.7% of the 1.8 million records in the sample.

Our dependent variable was the duration from a patient's assessment by a clinician to their departure from the ED. The distribution of this variable is heavily right-skewed. Thus, if we choose to perform a linear regression, we cannot expect residuals to be normally distributed or display constant variance (heteroscedasticity). Without action, this would affect the calculation standard errors and confidence intervals, though point estimates from such a model remain unbiased.

We used a heteroscedasticity-consistent (HC) estimator recommended by Long and Ervin ([2020](https://www.tandfonline.com/doi/abs/10.1080/00031305.2000.10474549)), HC3, to estimate the covariance of the model parameters, and obtain robust estimates of standard errors and confidence intervals.

GAM with Number of tests:

```{r}
#| eval: false

mgcv::gam(
  formula = dur_assess_depart ~
    # CASE-MIX-RELATED:
    s(age, by = sex) + sex + 
    arr_mode + acuity + chief_comp_desc + refer_sorc + # DIAGNOSIS 01 desc # N_dIAG
    # TIME-RELATED:
    month + wkday + hour + 
    # PROVIDER (RANDOM INTERCEPT):
    s(procode, bs = "re") +
    # VARS OF INTEREST:
    invst_50 + invst_51 + invst_52 + invst_53 + ...,
  family = "gaussian",
  method = "REML",
  data = df_prep_duration_integer
)

```
:::

```{r}
#| label: fig-test-time
#| echo: false
#| message: false
#| warning: false
#| column: page
#| fig-width: 10
#| fig-asp: 0.618
#| out-width: "85%"
#| fig-align: center

readRDS(here("data", "df_preplot_duration.rds")) |> 
  # REDUCE TO TOP 32 INVESTIGATIONS:
  semi_join(df_odds_results, join_by(InvestigationDescription)) |> 
    ggplot(aes(reorder(InvestigationDescription, dura), dura)) +
  geom_col(aes(reorder(InvestigationDescription, dura), dura), fill = "dodgerblue", alpha = .5)+ # , position = "dodge"
  geom_errorbar(aes(ymin = lci, ymax = uci), col = "grey40", alpha = .5, width = .3)+
  # geom_errorbar(aes(ymin = lciold, ymax = uciold), colour = "grey")+
  theme_minimal()+
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 9),
    legend.position = "none"
  )+
  labs(y = "\nEstimated (net) time added to a patient's attendance in 2023/24 (mins)")+
  coord_flip()
```

## A5. How does time in ED change according to the number of tests conducted?

As the number of tests per attendance has increased, so has the number of patients who receive several tests. In this section we examine the relationship between the number of tests that a patient receives and their duration in the ED. We describe this relationship in 2019/20 and in 2023/24. Unlike previous sections, this analysis is not casemix-adjusted, since the methods that are usually employed to adjust for casemix are unlikely to be adequate in this context (**perhaps add further note in an appendix?**). 

**A5: Results**

@fig-time-vs-tests shows that, in general, a patient’s (unadjusted) duration in ED tends to increase with the number of tests performed. This relationship can be explained in part with reference to patient casemix (more complex patients have more tests and stay longer), but also because, as we have seen above, most tests add time to a patient’s stay. 

However, the nature of this relationship has changed between 2019/20 and 2023/24. In 2019/20, 28 minutes was added to a patients stay on average, for each of the first 5 tests performed.fig-time-vs-tests But the impact of each additional test beyond the fifth was limited, adding on average only 7 minutes per test. By 2023/24, ED durations had increased considerably. Each of the first five tests now added 44 minutes on average. Moreover, for each additional test beyond the fifth, the time in ED continued to rise considerably - by 38 minutes per test. This change can be seen whether the patient was admitted or discharged, and whether the patient was treated for an injury or an illness (although subtle differences in the relationship are present (see appendix)).

Whilst this finding is clear in the data, it is not easy to explain. One potential line of enquiry would focus on the extent of asynchronous testing. The result might also be an emergent property of a system operating with reduced flow, which manifests as re-testing of patients whose care is delayed.

**Median time vs. number of tests**

```{r}
#| label: fig-time-vs-tests
#| echo: false
#| message: false
#| warning: false
#| column: page
#| fig-width: 10
#| fig-asp: 0.618
#| fig-align: center
#| out-width: "85%"


readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[5]] |>
  ggplot() +
  theme_minimal() +
  theme(
    axis.ticks = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  geom_line(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear
  )) +
  geom_point(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear,
    size = n_patients
  )) +
  scale_y_continuous(
    name = "Median time in ED (assessment to departure)",
    limits = c(0, NA_real_)
  ) +
  scale_x_continuous(
    name = "Number of tests",
    limits = c(0, 12),
    breaks = (0:12),
    labels = c(
      "0", "1", "2", "3", "4",
      "5", "6", "7", "8", "9",
      "10", "11", "12+"
    )
  ) +
  scale_size_continuous(name = "Number of attendances") +
  scale_colour_discrete(name = "Financial year") +
  labs(
    title = "Median time in ED by number of tests",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )

```

## A6. How do tests influence decisions to admit?

In this section, we explore how increased testing in ED has influenced the rate of admissions to hospital beds. We do this by comparing the change in the casemix-adjusted odds of a *test* with the casemix-adjusted odds of *admission* *having received a test*. We examine these changes over the period from 2019/20 to 2023/24, for 32 test types.

::: callout-note
## A6: Methods and Limitations

Methods and limitations here.
:::

@fig-adm-decision assigns each of the 32 tests to a quadrant, based on 2 values:

-   the change in the casemix-adjusted odds of a test, and

-   the change in the casemix-adjusted odds of admission having received a test.

Many of the tests - including most of the high-volume tests - fall into the bottom-left quadrant. This indicates that, after adjusting for casemix, patients were more likely to receive the test in 2023/24, but less likely to be admitted having received it. This is true for complete blood count, blood coagulation, c-reactive protein measurement, CT scan, hepatic function, and urea and electrolytes. 

Given that this analysis is casemix-adjusted, we can safely exclude casemix as an explanation for this finding. Thus, another theory is that clinicians are extending the use of tests as a means of ruling out conditions that would warrant admission. Unfortunately, this analysis cannot shed light on the motivation for such a change in practice. It might be that ED clinicians are proactively lowering test thresholds in response to changes in medical guidelines and emerging research. It might also be that test thresholds have fallen as tests have become more available, more timely, and less expensive. A further explanation might be that this change is driven by changes in bed availability. That is to say that, whilst, in the past, some tests may have been carried out after admission to a hospital bed, it may *now* be that these tests are performed in the ED. Further work would be required to determine the underlying cause of changes in testing practice. **(link to further work section?)**

Finally, there are two low-volume tests that appear to be outliers in this analysis: wound microscopy culture and sensitivities; and lipid panel. These appear in the top-right quadrant, but some way from the origin. Their position indicates that patients were much more likely to receive these tests in 2023/24, and much more likely to be admitted having received the test. This result is **subject to high levels of statistical uncertainty (?)** because the number of these tests that were performed in the baseline year (2019/20) was low. However, at face value, it suggests that these two tests are now used to identify patients who require admission. In 2019/20, these same patients would have been not have been admitted to a hospital bed.

```{r}
#| label: fig-adm-decision
#| echo: false
#| message: false
#| warning: false
#| column: page
#| fig-width: 10
#| fig-asp: 0.618
#| fig-align: center
#| out-width: "85%"

tmp_admodds <-
  readRDS(
  here("data", "from_ncdr_241106_admission_odds.rds")
  ) |> 
  filter(str_detect(term, "^fyear2023/24:")) |>
  # print(n=330)
  mutate(term = str_remove_all(term,"/")) |>
  mutate(term = str_remove(term,"fyear202324")) |> 
  mutate(term = str_remove_all(term,":")) |>
  mutate(term = str_sub(term, end = -2)) |> 
  select(term, odds_adm = odds)
  
# list.files(here("data_raw"))
  
lkp_invst_ec_aea <- readRDS(here("data_raw", "lkp_test_ec_to_aea.rds"))

df_odds_results |>
      filter(InvestigationDescription != "Serologic test") |>
  # NUMBER OF ATTENDANCES FROM ODDS MODEL DATA FRAME:
  mutate(p = tests_2023_24 /0.566e6) |>
  left_join(tmp_admodds, join_by(name == term)) |>
  select(term = name, InvestigationDescription, odds, odds_adm, id, starts_with("tests"), p) |>
  mutate(InvestigationKey = as.numeric(str_extract(term, "[:digit:]{2}"))) |>
  left_join(
    lkp_invst_ec_aea |> 
      select(InvestigationKey, invst_group), 
    join_by(InvestigationKey)
  ) |> 
  # count(invst_group) |> 
  ggplot()+
  geom_hline(yintercept = 1, lty = "dashed", alpha = 0.1)+ # col = "grey40")+
  geom_vline(xintercept = 1, lty = "dashed", alpha = 0.1)+ # col = "grey40")+
  geom_point(aes(odds, odds_adm, size = p), alpha = 0.3, stroke = NA)+
  scale_size(labels = scales::percent_format())+
  geom_text_repel(
    aes(
      odds, odds_adm,
      label = str_wrap(InvestigationDescription, 20)
      ),
    size = 2,
  ) +
  facet_wrap(vars(invst_group))+
  theme_bw()+
  theme(
    axis.ticks = element_blank(),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7),
    axis.title = element_text(size = 7)
  )+
  labs(
    x = "\nCase-mix-adjusted odds of test (2023/24 vs 2019/20)",
    y = "\nCase-mix-adjusted odds of admission having received test (2023/24 vs 2019/20)\n",
    size = "Attendances\nwith test\n2023/24"
  )
```

# Discussion

[**N - SW**]{.underline}

# Possible Further Research

# Appendix 1 {#sec-appendix}

[**O - AJ**]{.underline}

Make all sections of the appendix collapsable?

Sampling and SQL queries ? Collapsable?

<!-- To carry out these analyses, we require only the EC data set, and only two years (provisionally, 2019/20 and 2023/24). We intend to use different samples to answer the different types of question). We will include a provider in a sample if the quality of their data submissions meet our requirements for the given question. We expect this approach to  -->

### Sampling for Q1 (from trends dq)

Given that we will be covering a relatively long period, spanning two data sets (with slightly different structures), it will be important to understand the quality and consistency of recording practices, both over time and by provider. In earlier [work in this area](https://www.strategyunitwm.nhs.uk/publications/waiting-times-and-attendance-durations-english-ae-departments), we based our analysis on subsets of providers who had a history of high quality data. For this analysis, it is likely that we will, again, have to work with data from a subset providers.

There is more - will have to look...

We identified XXX providers with provider codes that survived across the 12 year study period and that, in addition, passed basic data quality checks. These XXX providers represent the total population of providers that we might have used for our analysis had data quality been universally high.

Providers were included in our sample if their data submissions met the quality required for this question. Here, the study period is relatively long but we have few variables to examine. We are looking to see that:

1.  Throughout the study period, providers have had consistently high levels of recording of the investigation fields.

2.  For attendances in which multiple investigations of the same type have been carried out, providers have recorded each instance of the investigation (and have been consistent in this practice over the whole study period).

We established that XX providers met these criteria. The providers we intend to use for our sample are given in @tbl-sample-q1. In the adjacent tab, @tbl-representation suggests that this sample of XXX providers is widely representative of the population of providers. We note that the sample includes higher proportions of larger providers and providers in London, and lower proportions of smaller providers and providers in the East of England.

## Sampling for Q2 (from ECDS DATA qual qmd)

One solution to this problem may be to use other variables to understand casemix (and may even be more valid measures). For example, the combination of *acuity* (composed of 5 levels) and Chief Complaint Group (15 groups) should allow us to account for differences in casemix. Both variables have \<13% NULL values in both years when looking at the data set as a whole. The completion rates would likely improve further if we were to examine a subset of providers.

It is possible, then, that the quality of a provider’s submission of these two variables, along with *Der_EC_Investigation_All*, may determine whether we include the provider in our analyses. We will address this idea next.

we deduce that 16 providers can claim to have **all 3 fields** **complete** for \>90% of attendances, in both years. If we were to relax the threshold to \>85%, we get just four more providers.

### How representative is the provider sample?

We identified 102 providers with provider codes that were present in the two years of interest and that, in addition, passed basic data quality checks. These providers will represent the total population of providers that we might have used for our analyses. Given the data quality issues, we have selected a sample of providers to improve the reliability and validity of our results. However, since we wish to make general statements about about the effect of diagnostic testing on urgent and emergency care, we must ensure that our samples are representative of the population of providers.

More deprived, and older population, with more arrivals by ambulance results in more admissions. We will control for all of these factors - what difference does this make…?

# Appendix 2

::: {.callout-note collapse="true" appearance="minimal" icon="false"}
## title

Additional graphics relating to A5.

```{r}
#| label: fig-tests-per-attd
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[1]] |>
  ggplot() +
  theme_bw() +
  # theme_minimal() +
  geom_vline(aes(xintercept = mean_tests, colour = fyear), linetype = "dashed") +
  geom_line(aes(x = n_tests_all, y = p_patients, colour = fyear)) +
  geom_point(aes(x = n_tests_all, y = p_patients, colour = fyear)) +
  scale_y_continuous(
    name = "Proportion of attendances",
    label = label_percent(accuracy = 1)
  ) +
  scale_x_continuous(name = "Number of tests") +
  labs(
    # title = "Number of tests per attendance",
    # subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24",
    caption = "Selected providers | Apr19-Feb20 & Apr23-Feb24\nDashed lines indicate mean number of tests"
  )
```

```{r}
#| label: fig-grpd-tests-per-attd
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[2]] |>
  ggplot() +
  geom_col(
    aes(fyear, p_patients_adj, fill = n_tests_all_grp),
    position = position_stack()
  ) +
  geom_hline(aes(yintercept = 0), colour = "grey20", lty = "dashed") +
  geom_text(aes(x = fyear, y = p_patients_adj, group = n_tests_all_grp, label = p_patients_label),
    position = position_stack(vjust = 0.5)
  ) +
  # scale_fill_manual(values = c("red", "orange", "yellow", "grey")) +
  # scale_colour_manual(values = c("red", "orange", "yellow", "grey")) +
  scale_y_continuous(
    name = "Proportion of attendances",
    label = label_percent(accuracy = 1)
  ) +
  scale_x_discrete(name = "Financial year") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.y = element_blank()
  ) +
  labs(
    # title = "Number of tests per attendance",
    # subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24", 
    caption = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )
```
:::

::: {.callout-note collapse="true" appearance="minimal" icon="false"}
# Graphics (click to unhide)

**Tests per attendance**

```{r}

```

**Tests per attendance (grouped)**

```{r}

```

### Grouped by presentation and disposal

```{r}
#| label: fig-grpd-tests-per-attd-disp
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[3]] |>
  ggplot() +
  geom_col(aes(x = fyear, y = p_patients_adj, fill = n_tests_all_grp),
    position = position_stack()
  ) +
  geom_hline(aes(yintercept = 0), colour = "grey20", lty = "dashed") +
  geom_text(
    aes(x = fyear, y = p_patients_adj, group = n_tests_all_grp, label = p_patients_label),
    position = position_stack(vjust = 0.5),
    size = 2
  ) +
  facet_grid(cols = vars(is_admitted), rows = vars(inj_illness)) +
  scale_fill_manual(values = c("red", "orange", "yellow", "grey")) +
  scale_y_continuous(
    name = "Proportion of attendances",
    label = label_percent(accuracy = 1)
  ) +
  scale_x_discrete(name = "Financial year") +
  theme(legend.title = element_blank()) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Number of tests per patient by disposal and presentation type",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )
```

**Grouped by test type**

```{r}
#| label: fig-grpd-tests-per-attd-type
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[4]] |>
  ggplot() +
  geom_col(aes(x = fyear, y = p_patients_adj, fill = n_tests_grp),
    position = position_stack()
  ) +
  geom_hline(aes(yintercept = 0), colour = "grey20", lty = "dashed") +
  geom_text(aes(x = fyear, y = p_patients_adj, group = n_tests_grp, label = p_patients_label),
    position = position_stack(vjust = 0.5)
  ) +
  facet_wrap(vars(test_type)) +
  scale_fill_manual(values = c("orange", "yellow", "grey")) +
  scale_y_continuous(
    name = "Proportion of attendances",
    # label = label_percent(accuracy = 1),
    breaks = c(-0.5, -0.25, 0, 0.25, 0.5),
    labels = c("50%", "25%", "0%", "25%", "50%")
  ) +
  scale_x_discrete(name = "Financial year") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Number of tests per patient by test type",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )

```

### Time by presentation and disposal

```{r}
#| label: fig-time-vs-tests-grp-disp
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[6]] |>
  ggplot() +
  # theme_minimal()+
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  geom_line(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear
  )) +
  geom_point(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear,
    size = n_patients
  )) +
  facet_grid(
    cols = vars(is_admitted),
    rows = vars(inj_illness)
  ) +
  scale_y_continuous(
    name = "Median time in ED (assessment to departure)",
    limits = c(0, NA_real_)
  ) +
  scale_x_continuous(
    name = "Number of tests",
    limits = c(0, 12),
    breaks = (0:12),
    labels = c(
      "0", "1", "2", "3", "4",
      "5", "6", "7", "8", "9",
      "10", "11", "12+"
    )
  ) +
  scale_size_continuous(name = "Number of attendances") +
  scale_colour_discrete(name = "Financial year") +
  labs(
    title = "Median time in ED and number of tests by dispsosal and presentation type",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )

```

### Time by test type

```{r}
#| label: fig-time-vs-tests-grp-type
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[7]] |>
  ggplot() +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  geom_line(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear
  )) +
  geom_point(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear,
    size = n_patients
  )) +
  facet_wrap(vars(test_type)) +
  scale_y_continuous(
    name = "Median time in ED (assessment to departure)",
    limits = c(0, NA_real_)
  ) +
  scale_x_continuous(
    name = "Number of tests",
    limits = c(0, 5),
    breaks = (0:5),
    labels = c(
      "0", "1", "2", "3", "4",
      "5+"
    )
  ) +
  scale_size_continuous(name = "Number of attendances") +
  scale_colour_discrete(name = "Financial year") +
  labs(
    title = "Median time and number of tests in ED by test type",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )


```
:::
