---
title: "Growth in Diagnostic Testing and Its Impact on Emergency Departments"
#"Growth in Diagnostic Testing<br>and Its Impact on<br>Emergency Departments"
subtitle: "<br>Steven Wyatt and Andrew Jones | The Strategy Unit | November 2024"
# author: 
#   - name: Steven Wyatt
#     email: swyatt@nhs.net
#     corresponding: TRUE
#     affiliation: The Strategy Unit
#   - name: Andrew Jones
#     email: swyatt@nhs.net
#     corresponding: FALSE
# date: 2024-11-29
format:
  html:
    html-table-processing: none
    # code-fold: true
    code-tools: true
    embed-resources: true
    grid:
      margin-width: 280px
  pdf: default
reference-location: margin
citation-location: margin
toc: true
toc-title: Contents
toc-location: left
theme: journal
fig-cap-location: top
number-sections: false
title-block-banner: '#f8be06'
title-block-banner-color: '#242322'
# mainfont: Gill Sans MT
# mainfont: Helvetica
editor_options: 
  chunk_output_type: console
---

*Steven:*

*I've been focussing on the parts that are likely more easily done by me. This includes insertion of the graphics/tables but also the necessary, but less-likely-to-be-read parts, e.g. Methods, Limitations).*

*I thought you'd be better placed to:*

*- write the analysis for questions I haven't completed / edit the existing text as you see fit, and*

*- write/edit the more subjective parts (e.g. Discussion, Exec Sum, framing of the Intro).*

*Note that the additional work you did on multiple tests, I've labelled as Q5. And what was formerly Q5 is now Q6.*

# Executive Summary

[**A - SW**]{.underline}

Notes: Andrew Jackson's suggested/preferred template:

Last time we said this, now we're saying this. (ie. where is the new work consistent with past work - where different.)

\[He is intending to use this as further evidence to suggests that 4HR target is not appropriate\]

# Introduction

[**B - SW**]{.underline}

Over the last decade, growth in diagnostic testing in emergency departments (EDs) has been rapid. Earlier analysis conducted by the Strategy Unit ([2019](https://www.strategyunitwm.nhs.uk/publications/waiting-times-and-attendance-durations-english-ae-departments), [2023](https://www.strategyunitwm.nhs.uk/sites/default/files/2023-05/MDSN%20Diagnostic_growth_report%20Apr2023.pdf)) suggests that this growth may have implications for the timeliness of patient care, for decisions on whether to admit a patient, and for crowding in EDs.

::: column-margin
For the sake of readability, we use the term, “test”, to refer to a unit of diagnostic activity. We acknowledge that some of these “tests” might also be therapeutic procedures.
:::

In this report, we update selected analyses from these past reports , in order to... *\[how do we wish to frame this? - Alternatively explain after box below\]*

We seek answers to six questions:

::: {.callout-note appearance="minimal" icon="false"}
#### Questions on diagnostic testing in EDs that we address in this report:

<br>

1.  How has the number of tests per attendance changed over time?

2.  Can the rise in test rates be explained by changes to casemix?

3.  What is driving the growth in testing?

    <!--(Is it changes to demand, casemix, or practice?)  -->

4.  How much time do tests add to a patient’s attendance?

5.  How does time in ED change according to the number of tests conducted?

6.  How do tests influence decisions to admit?
:::

It is hoped that answers to these questions .... \[how do we wish to frame this?\]

<!--In the second phase of this project, for which findings are due in the spring of 2025, we will disregard the boundary between EDs and inpatient beds. And look at the impact of diagnostic growth on urgent and emergency care a inpatient beds. Rather than looking at the im... might ///Increased diagnostic testing means that patients spend less time in contact with hospital services, even ifpatients an increase in time in ED.///we measure the total time that patients spend in contact with hospital services, starting when they enter the UEC pathway.-->

# Analytical Approach {#sec-approach}

[**C- AJ**]{.underline}

The analyses we present here investigate the consequences of increased diagnostic testing within major emergency (Type 1) departments. To address the questions above, we relied on SUS A&E and EC data sets.

Whilst these are high-quality, patient-level sources, we observed considerable variation in recording practices, both across the population of provider trusts and over time. If overlooked, this variation would adversely affect both the [reliability](https://en.wikipedia.org/wiki/Reliability_(statistics)) and [validity](https://en.wikipedia.org/wiki/Validity_(statistics)#Reliability) of our results.

#### The use of provider sampling {#sec-approach-sampling}

To increase the chances that our findings would be reliable and valid for a given question, we conducted each analysis on data from a sample of providers. These provider samples were non-random: we included a provider in a sample if the quality of their data submissions met our requirements for the given question.

Whilst we addressed six distinct questions, we were able to identify two different samples. We identified a sample of 28 providers with which we could address Question 1. A second sample, of 16 providers, could be used for the analysis of Questions 2 to 6.

Whilst sampling should improve the reliability and validity of our results, we cannot make general statements about the effect of testing on EDs *unless* our sample of providers is *representative* of the population of providers. That is to say, the providers in our sample(s) and population should have similar characteristics.

<!--With a representative sample, we can be more confident that our findings for the sample are also applicable to the population.-->

There is a trade-off to be made here. The larger the sample of providers we select, the more likely it will be that this

The more providers we capture in our sample, the more likely the sample is to be representative of the population of providers.

The larger our sample of providers, the more likely it is to be representative of the population of providers. But the larger we make our sample, the more we must compromise on data quality, which will likely to decrease the the reliability of our results.

Details of the sampling methods, our approach to the trade-off, and the measures of representation can be found in @sec-appendix.

#### Excluded attendances {#sec-approach-attd}

For providers within our samples, the analyses cover all unplanned attendances of persons with an English postcode, with some exceptions. We excluded cases were a person left the ED before being treated, and persons streamed to other care services. In addition, we excluded cases where the patient died during an attendance and those where the patient was found to be dead on arrival at the ED.

#### Question-specific approaches

Details of our approach to each distinct question can be found in the expandable "Methods and Limitations" boxes, in the corresponding subsections of the Analysis section, below.

#### Reproducibility

The computer code used to perform this analysis can be found a [GitHub repository](https://github.com/The-Strategy-Unit/diagnostic_growth_uec).

# Analysis

```{r}
#| label: setup
#| include: false

library("gt")
library("here")
library("dplyr")
library("tidyr")
library("scales")
library("forcats")
library("ggrepel")
# library("purrr")
# library("readr")
library("ggplot2")
library("janitor")
library("stringr")
library("gtExtras")
library("lubridate")
```

#### Perhaps some context - if time?

**D - AJ/SW (depending on time available)**

e.g. Describe the pop attending EDs show the casemix in both years, age, condtion... can't exclude all casemix effects

## A1. How has the number of tests per ED attendance changed over time?

::: {#nte-q1-meth .callout-note collapse="true" appearance="minimal" icon="false"}
# A1: Method and Limitations (click to unhide)

<br>**Method**

To be more precise, we estimated how the *average* number of tests per ED attendance has changed over time. As this analysis covers a period of 12 years - starting with the financial year 2012/13 and concluding with 2023/24 - we had to employ both A&E and EC data sets.

We identified 28 providers with provider codes that survived across the 12 year study period and that passed our data quality criteria. See @sec-appendix.

The numerator for our "test rate" in a given year was the total number of tests performed across our provider sample. Where a patient received multiple tests of the same type, we counted these as multiple tests. The denominator for our "test rate" in a given year was the total number of ED attendances across our provider sample that matched the criteria given in @sec-approach-attd.

**Limitations**

Our aim is to make general statements about the effect of diagnostic testing on urgent and emergency care. Due to concerns over data quality, we employed sampling in this analysis. Our results come from 28 of 99 providers whose provider codes remained unchanged over the 12-year period. As the 28 providers were chosen on the basis of their data quality, this constitutes a non-random sample of providers. Study conclusions drawn from non-random samples are at higher risk of bias. As noted in @sec-approach-sampling, we attempted to mitigate the risk of bias by confirming that the sample was generally representative of the population of providers in various dimensions. It is, however, possible that we unknowingly omitted important dimensions for this particular question, and that our findings are less generalisable than we suggest.

There are two additional points worth noting. Firstly, in order to complete this analysis, we were obliged to unite A&E and EC data sets, each of which has a slightly different structure. We used the A&E data set for financial years between 2012/13 and 2018/19 and the EC data set between 2019/20 and 2023/24. Whilst we attempted to ensure consistency across data sets, it may be that there are some differences

The second important point is that we included only those providers who appeared to record multiple instances of distinct investigations (and who appeared to have been consistent in this practice over the whole study period). It is, however, possible that these providers were internally inconsistent in recording practices over the period or that, in some cases, the recording of multiple tests of the same type represented an error. \[Could reference our data analysis here\]
:::

**Analysis 1: Results**

[**E - SW (to edit as necessary)**]{.underline}

@fig-trend shows the change in test rates for the sample of 28 providers identified above. It appears that, overall (as shown in the left-hand panel), test rates doubled over the study period, from an average of 2 tests per attendance in 2012/13 to 4 tests per attendance in 2023/24.

For the 3 in 10 patients who are are admitted after an ED attendance, the recorded rate of increase was uneven but appears to have slowed over the period. During the pandemic period, ED attendances generally fell, while evidence suggests that persons who did attend ED were sicker. We suggest that the trend is now normalising after the short-term spike in test rates above the trend.

For the 7 in 10 patients who are are not admitted after an ED attendance, the increase was fairly consistent to 2020. Again, the rate of increase seems to have slowed following the spike in test rates (seen across the majority of test types) in 2020/21.

<br>

::: panel-tabset
#### Graphical Presentation

```{r}
#| label: fig-trend
#| echo: false
#| fig-cap: "Estimated average number of tests per ED attendance, in England, between 2012/13 and 2023/24. Source: SUS A&E and EC data sets."
#| message: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

df_preplot_trend <- readRDS(
  here("data_raw", "from_ncdr_trends_241003_df_preplot.rds")
  ) |> 
  mutate(disdest = fct_relevel(disdest, "overall (all destinations)")) |> 
  mutate(disdest = fct_relevel(disdest, "non-admitted patients", after = 1))

# FOR GRAPHIC FYEAR AXIS LABELS:
every_nth <-  function(n) {
  return(function(x) {x[c(TRUE, rep(FALSE, n - 1))]})
}

df_preplot_trend |>
  ggplot(aes(fyear, rate)) +
  geom_line(aes(col = disdest, group = disdest)) +
  geom_point(aes(col = disdest, group = disdest)) +
  geom_blank(aes(y = 0)) +
  scale_color_manual(values = c("indianred3", "dodgerblue", "purple4")) +
  labs(x = "Financial year")+
  theme_bw()+
  labs(y = "Average number of recorded tests per ED attendance")+
  theme(
    legend.position = "none",
    axis.ticks = element_blank()
    )+
  facet_wrap(vars(disdest))+ # , scales = "free_y"
  scale_x_discrete(breaks = every_nth(n=4))+
  scale_y_continuous(breaks = seq(0,9, by = 2))
```

#### Tabular Presentation

If we want numerators/denoms, we'll need to reproduce the source object for the graph, which means going back to NCDR
:::

## A2. Can the rise in test rates be explained by changes to casemix?

[**F - SW (intro?)**]{.underline}

<!--Start off with some intro text from Diagnostics report 2023?  -->

<!--Previous work by the Strategy Unit suggests that people seeking medical attention in ED today are sicker than those arriving in the past. We addressed this question for 32.  To answer this question, we looked at whether patients are more likely to receive tests in EDs today than they were in 2019/20. After adjusting for casemix, are patients more likely to receive tests in EDs today than they were in 2019/20? . Having adjusted for casemix, how much more likely it is that a patient will receive a test in an ED in 2023/24 compared to 2019/20?  -->

::: {#nte-q2-meth .callout-note collapse="true" appearance="minimal" icon="false"}
# A2: Method and Limitations (click to unhide)

<br>**Method**

We estimated the casemix-adjusted odds of patient receiving a test in an ED in 2023/24, when compared to 2019/20.

We used the SUS EC data set and identified 16 providers with provider codes that survived across the study period and that passed our data quality criteria for this question. For further details of our sampling approach and data quality criteria see @sec-approach-sampling and @sec-appendix.

We built 32 distinct models to predict casemix-adjusted odds ratios (as specified above) for 32 (of 49) test types that appear in the EC data set. These 32 tests account for over 99% of all recorded tests in EDs.

In terms of data preparation, we removed March from both 2019/20 and 2023/24, to avoid any bias to our results that may have arisen due to the onset of the pandemic in March 2020.

Approximately 4% of the 3 million records had missing values in a key field (our "key fields" are given in @sec-appendix). Our specified model type will not accept missing values, so we chose to delete the records with missing values in our key fields.

We performed logistic regressions using generalised additive models (GAMs) with mixed-effects. All 32 models were specified in the same way, with the only difference being the binary dependent variable (whether the given test was carried out). Our fixed effects included the variable of interest, financial year, and several time-related variables to control for season, weekday, and time of arrival. We included variables to adjust for casemix (as far as possible): *age* (as a spline-based smooth) interacting with *sex*; a re-engineered *chief complaint* variable; *acuity*; *arrival mode*; and *referral source*.

(It is important to note that the *chief complaint* and *acuity* variables are recorded as part of an initial assessement ([NHSE, 2022](https://www.england.nhs.uk/guidance-for-emergency-departments-initial-assessment/#:~:text=chief%20complaint%20or%20suspected%20diagnosis,initially%20assessed%20by%20a%20clinician)). They are akin to presentation codes. Whilst the diagnosis codes may give greater detail about a patient's condition, diagnosis is often a causal consequence of testing and it would therefore be inappropriate to use diagnosis-related information to predict whether a patient receives a given test.)

We treated the provider variable as a random effect. This allowed us to control for factors that vary, at random, across providers. Moreover, the values produced by this random effect can be thought of as sample values from a larger population of values. As the providers in our data set were, in fact, a sample of all providers, this approach suits our case well.

We ran each model on the same sample of 1 million attendance records.

**Limitations**

The findings of this analysis are based on a sample of 16 providers (of 102 providers whose provider codes remained unchanged over the study period). These findings are therefore subject to the same sampling-related limitations we identified for Q1 (see @nte-q1-meth).

We have labelled our estimates as "casemix-adjusted" as we controlled for a range of variables that might indicate a patient's need for a test. However, the clinician who assesses a patient will have far more information on that patient's condition than we have access to.

the fact th the fact that a clinician requested (or did not request) a test is likely to be indicative of a patient's condition and because we don't Because we don't have access to all the information used by the clinician, we can only partially control for casemix.

As we are missing, the fact that a clinician is likely to be indicative of a patient's condition. There may be many more variables that influence whether a test is given than we have access to. and the fact that a clinician requested (or did not request) a test, itself, is also likely to be indicative of a patient's condition. Because we are unable to control for unrecorded variables, we can only partially control for casemix, and it's possible that more (or less) of the change we've seen is due to casemix.

when deciding whether or not to order a test for a patients, clinician

We did not take into account the fact that a clinician requested (or did not request) a test. Because we are unable to control for these unrecorded variables, we cannot fully adjust our estimates for casemix, and our estimates may be biased. it's possible that more (or less) of the change we've seen is due to casemix.

and whilst it is true to say that

It is true to say this may only be a partial adjustment for casemix.

Clinicians have information, when deciding whether to order a test for a patients. Because we are unable to control for these unrecorded variables, we cannot fully adjust our estimates for casemix, and our estimates may be biased. it's possible that more (or less) of the change we've seen is due to casemix.

This information is also likely to be indicative of a patient’s condition. There is some that cannot be controlled for.

Finally, as noted in the Methods section, above, we decided to delete approximately 4% of records from our sample that had missing values for any of the variables used in the model. We expect this decision to have a limited impact on our results. However, as it reduced our sample size, the standard errors around our model coefficients have marginally increased as a consequence. There is also the potential for some bias in our results if these missing data were not missing completely at random.
:::

**Analysis 2: Results**

[**G - SW**]{.underline}

A description and analysis of the the graphic(s) below.

::: panel-tabset
## Graphic<br>(Excl. serological testing)

```{r}
#| label: fig-odds-sans-sero
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

df_odds_results <- bind_rows(
  readRDS(here("data", "from_ncdr_growth_attrb_v3_1to16.rds")),
  readRDS(here("data", "from_ncdr_growth_attrb_v3_17to32.rds"))
) |>
  # THE NUMBER OF 2023/24 ATTENDANCES IN NESTED DFS FOR ODDS MODEL:
  mutate(att_2023_24 = 0.566e6) |>
  mutate(p = tests_2023_24 / att_2023_24)


# df_odds_results |>
  # filter(InvestigationDescription != "Serologic test") |> 
  # ggplot() +
  # geom_hline(yintercept = 1, lty = "dashed", colour = "grey20") +
  ## geom_hline(yintercept = 2, lty = "dashed", colour = "grey20") +
  # geom_col(aes(reorder(InvestigationDescription, odds), odds), alpha = .8) +
  # theme_bw() +
  # theme_minimal() +
  # coord_flip() +
  # scale_y_continuous(
  #   breaks = seq(0, 4, by = 1),
  #   labels = c(
  #     "",
  #     # "0.5x\nas likely in\n2023/24",
  #     "Equally\nas likely in\n2023/24",
  #     # "1.5x\nas likely in\n2023/24",
  #     "2x\nas likely in\n2023/24",
  #     "3x\nas likely in\n2023/24",
  #     "4x\nas likely in\n2023/24"
  #     ),
  #   position = "right"
  # ) +
  # theme(
  #   axis.title.y = element_blank(),
  #   plot.margin = margin(5, 20, 5, 5),
  #   panel.grid.minor = element_blank()
  # ) +
  # labs(y = "Case-mix-adjusted odds ratio of test (2023/24 vs 2019/20)\n")


df_odds_results |>
    filter(InvestigationDescription != "Serologic test") |>
  ggplot(aes(odds, p)) +
  geom_vline(xintercept = 1, lty = "dashed", colour = "grey20") +
  # theme_bw() +
  theme_minimal() +
  geom_point() +
  geom_blank(aes(x = 0, y = 0)) +
  geom_text_repel(
    aes(label = str_wrap(InvestigationDescription, 20)),
    size = 2,
  ) +
  scale_y_continuous(
    limits = c(0, 0.5),
    labels = scales::percent_format()
  ) +
  scale_x_continuous(
    breaks = seq(0, 4, by = 1),
    labels = c(
      "",
      # "0.5x\nas likely in\n2023/24",
      "Equally as likely\nin 2023/24",
      # "1.5x\nas likely in\n2023/24",
      "2x as likely\nin 2023/24",
      "3x as likely\nin 2023/24",
      "4x as likely\nin 2023/24"
      ),
    position = "top"
  ) +
  labs(
    x = "Case-mix-adjusted odds ratio of test (2023/24 vs 2019/20)\n",
    y = "Proportion of attendances involving test (2023/24)\n"
  )+
  theme(
        plot.margin = margin(5, 20, 5, 5)
  )+
  NULL

```

## Graphic<br>(Incl. serological testing)

```{r}
#| label: fig-odds
#| echo: false
#| message: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "80%"
#| out-height: "80%"

df_odds_results |>
  # filter(InvestigationDescription != "Serologic test") |> 
  ggplot() +
  geom_hline(yintercept = 1, lty = "dashed", colour = "grey20") +
  # geom_hline(yintercept = 2, lty = "dashed", colour = "grey20") +
  geom_col(aes(reorder(InvestigationDescription, odds), odds), alpha = .8) +
  theme_bw() +
  theme_minimal() +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 5),
    labels = c(
      "",
      # "0.5x\nas likely in\n2023/24",
      # "Equally\nas likely in\n2023/24",
      # "1.5x\nas likely in\n2023/24",
      "5x\nas likely in\n2023/24",
      "10x\nas likely in\n2023/24",
      "15x\nas likely in\n2023/24",
      "20x\nas likely in\n2023/24"
      ),
    position = "right"
  ) +
  theme(
    axis.title.y = element_blank(),
    plot.margin = margin(5, 20, 5, 5),
    panel.grid.minor = element_blank()
  ) +
  labs(y = "Case-mix-adjusted odds ratio of test (2023/24 vs 2019/20)\n")
```

## Table

```{r}
#| label: tab-odds
#| echo: false
##| fig-cap: "Having adjusted for casemix, how much more likely it is that a patient will receive a test in an ED in 2023/24 compared to 2019/20? Source: ECDS."
#| message: false
##| fig-align: center
#| column: page
##| fig-width: 8
##| fig-asp: 0.618
##| out-width: "90%"

df_odds_results |>
  select(InvestigationDescription, att_2023_24, tests_2023_24, odds) |>
  mutate(tests_2023_24 = as.integer(tests_2023_24)) |> 
  gt() |>
  cols_label(
    InvestigationDescription = "Test name",
    att_2023_24 = "Attendances in sample<br>2023/24",
    tests_2023_24 = "Tests in sample<br>2023/24",
    odds = "Odds ratio of test (casemix adj.)<br>2023/24 vs. 2019/20", 
        .fn = md
  ) |> 
  fmt_auto() |>
  tab_options(
    data_row.padding = px(1),
    table.font.size = "small"
  ) 

```
:::

## A3. What is driving test growth in EDs?

[**H - SW (intro?)**]{.underline}

Intro text: Is the growth in EDs being driven by changes to demand, casemix, or practice?

::: {.callout-note collapse="true" appearance="minimal" icon="false"}
# A3: Method and Limitations (click to unhide)

<br>**Method**

For each of the 32 tests covered in Q2, we set out to find the percentage growth in attendances with one or more test ($T$), as a function of the growth due to: changes in raw demand ($D$), changes in casemix ($C$), and changes driven by all other factors ($P$). (We expect $P$ to be dominated by the influence of clinical practice, but it may also contain other factors, such as the availability of resources.)

Again, this analysis relies on the SUS EC data set and the 16 providers with provider codes that survived across the study period and that passed our data quality criteria for this question. For further details of our sampling approach and data quality criteria see @sec-approach-sampling and @sec-appendix.

To address the question, we formulated a multiplicative model

$1+T = (1+D)(1+C)(1+P)$ . \[1\]

Expanding the right hand side and solving for $T$ we get:

$T = D + C + [P+ DC + DP + CP + DCP]$ . \[2\]

Thus, we are able to isolate the main effects for growth due changes in demand and casemix. The terms within the square brackets include the $P$ term and composite, or interaction, terms (which are in most cases likely to be negligible).

We know that $T$ is,

$T = t_{23/24}/t_{19/20} -1$ , \[3\]

where $t$ is the number of attendances with one or more test in the specified financial year.

We might also define $D$, independently, as,

$D = a_{23/24}/a_{19/20} -1$ , \[4\]

where $a$ is the number of attendances (both with and without a test) in the financial year.

We then move to a calculation of $P$ which relies on predictions from our logistic regression models from Q2.

We first create a counterfactual scenario in which the casemix in the target year is replaced by the casemix in the baseline year attended in the target year. If we compare this counterfactual scenario for In this case, both $D$ and $C$ will be equal to zero, and Equation 2 becomes:

$T_{cf} = P$ . \[5\]

where

$T_{cf} = t_{23/34(cf)} / t_{19/20} -1$ . \[6\]

We can find $T_{cf}$ using predictions from the logistic regression models produced in Analysis 2. If we edit our data so We use the applicable model to predict

$t_{19/20}$ . (For our counterfactual scenario, $t_{23/34(cf)}$ , we make predictions having re-labelled the 2019/20 attendances as 2023/24 attendances.)

First we used the regression model to

We predicted the number of attendances with a test in the baseline year, divided by Then we predict the number of attendances with a test for a counterfactual scenario where the target year had the same casemix as the baseline year.

Expanding this we get,
:::

::: {.callout-note collapse="true" appearance="minimal" icon="false"}

**Limitations**

The findings of this analysis are based on a sample of 16 providers (of 102 providers whose provider codes remained unchanged over the study period). These findings are therefore subject to the same sampling-related limitations we identified for Q1 (see @nte-q1-meth).

Moreover, the analysis relies on the logistic regression models we built for Q2, hence the limitations we acknowledged in that analysis (see @nte-q2-meth) also apply here.
:::

**Analysis 3: Results**

[**I - SW**]{.underline}

A description and analysis of the the graphic below.

```{r}
#| include: false

df_preplot_growth_components <- df_odds_results |>
  mutate(d0 = 1, .before = dT) |>
  relocate(dC, .before = dP) |>
  pivot_longer(cols = matches("^d|^abs"), names_to = "component", values_to = "value") |>
  select(-tests_2023_24) |>
  mutate(value_abs = tests_2019_20 * value) |>
  mutate(value_order = if_else(component == "d0", value_abs, NA_real_)) |>
  mutate(value = if_else(component == "d0", NA_real_, value)) |>
  mutate(component = case_when(
    component == "d0" ~ "Attds with test\n19/20",
    component == "dT" ~ "Growth in attds with test\n19/20 to 23/24",
    component == "dD" ~ "Growth driven by demand",
    component == "dC" ~ "Growth driven by case-mix",
    component == "dP" ~ "Growth driven by practice",
    component == "dI" ~ "Growth due interactions",
    T ~ NA_character_
  )) |>
  mutate(component = forcats::as_factor(component)) |>
  mutate(value_abs = as.integer(value_abs)) |>
  group_by(InvestigationDescription) |>
  mutate(value_order = max(value_order, na.rm = T)) |>
  ungroup() |>
  mutate(value_abs = round(value_abs / 1000)) |>
  mutate(value_label_abs = if_else(is.na(value), str_c(value_abs, "k"), NA_character_)) |>
  mutate(value_label_p = if_else(value >= 0, str_c("+", round(value * 100), "%"), str_c(round(value * 100), "%")))


```

::: panel-tabset
## Percentage differences

```{r}
#| label: fig-attrib-perc
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-height: 10
#| out-width: "90%"

# #| fig-asp: 0.618

df_preplot_growth_components |>
  ggplot(aes(reorder(InvestigationDescription, value_order), value_abs)) +
  geom_col(
    aes(
      reorder(InvestigationDescription, value_order), value_abs,
      fill = component
    ),
    position = "identity",
    alpha = 0.4
  ) +
  geom_text(
    aes(
      label = value_label_abs,
      vjust = ifelse(value_abs >= 0, -0.4, 1.4),
      angle = 270
    ),
    size = 1.5,
    colour = "grey20"
  ) +
  geom_text(
    aes(
      label = value_label_p,
      vjust = ifelse(value_abs >= 0, -0.4, 1.4),
      angle = 270
    ),
    size = 1.5,
    colour = "grey20"
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, lty = "dashed") +
  theme_minimal() +
  theme(
    axis.ticks = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 5),
    panel.grid.minor = element_blank(),
    # panel.spacing = unit(1, "lines")
  ) +
  scale_y_continuous(
    limits = c(-50, 250)
  ) +
  facet_wrap(vars(component), ncol = 6) +
  labs(y = "\nED attendances (thousands)\nwith 1+ test of specified type")
```

## Absolute differences (for sample)

```{r}
#| label: fig-attrib-abs
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-height: 10
#| out-width: "90%"

# #| fig-asp: 0.618

df_preplot_growth_components |>
  mutate(value_label_abs = str_c(value_abs, "k")) |>
  ggplot(aes(reorder(InvestigationDescription, value_order), value_abs)) +
  geom_col(
    aes(reorder(InvestigationDescription, value_order), value_abs, fill = component),
    position = "identity",
    alpha = 0.4
  ) +
  geom_text(
    aes(
      label = value_label_abs,
      vjust = ifelse(value_abs >= 0, -0.4, 1.4),
      angle = 270
    ),
    size = 1.8,
    colour = "grey20"
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, lty = "dashed") +
  theme_minimal() +
  theme(
    axis.ticks = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 5),
    panel.grid.minor = element_blank(),
    # panel.spacing = unit(1, "lines")
  ) +
  scale_y_continuous(
    limits = c(-50, 250)
  ) +
  facet_wrap(vars(component), ncol = 6) +
  labs(y = "\nED attendances (thousands)\nwith 1+ test of specified type")


```

## Table

To come.
:::

## Q4. How much time do tests add to a patient's stay in the ED?

[**J - AJ**]{.underline} **(work in progress)**

How much (net) time does a test add to a patient’s stay in ED?

## Q5. How does time in ED change according to the number of tests conducted?

[**K - SW (text)**]{.underline}

**Results**

Text here.

**Median time vs. number of tests**

```{r}
#| label: fig-time-vs-tests
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"


readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[5]] |>
  ggplot() +
  theme_minimal() +
  theme(
    axis.ticks = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  geom_line(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear
  )) +
  geom_point(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear,
    size = n_patients
  )) +
  scale_y_continuous(
    name = "Median time in ED (assessment to departure)",
    limits = c(0, NA_real_)
  ) +
  scale_x_continuous(
    name = "Number of tests",
    limits = c(0, 12),
    breaks = (0:12),
    labels = c(
      "0", "1", "2", "3", "4",
      "5", "6", "7", "8", "9",
      "10", "11", "12+"
    )
  ) +
  scale_size_continuous(name = "Number of attendances") +
  scale_colour_discrete(name = "Financial year") +
  labs(
    title = "Median time in ED by number of tests",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )

```

::: panel-tabset
## 

```{r}

```

## 

```{r}

```

## 

```{r}


```

## 

```{r}


```

## 

```{r}

```

## 

```{r}

```

## 

```{}
```
:::

## Q6. How do tests influence decisions to admit?

[**L - SW (intro?)**]{.underline}

[**M - SW (results)**]{.underline}

```{r}
#| label: fig-adm-decision
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

tmp_admodds <-
  readRDS(
  here("data", "from_ncdr_241106_admission_odds.rds")
  ) |> 
  filter(str_detect(term, "^fyear2023/24:")) |>
  # print(n=330)
  mutate(term = str_remove_all(term,"/")) |>
  mutate(term = str_remove(term,"fyear202324")) |> 
  mutate(term = str_remove_all(term,":")) |>
  mutate(term = str_sub(term, end = -2)) |> 
  select(term, odds_adm = odds)
  
# list.files(here("data_raw"))
  
lkp_invst_ec_aea <- readRDS(here("data_raw", "lkp_test_ec_to_aea.rds"))

df_odds_results |>
      filter(InvestigationDescription != "Serologic test") |>
  # NUMBER OF ATTENDANCES FROM ODDS MODEL DATA FRAME:
  mutate(p = tests_2023_24 /0.566e6) |>
  left_join(tmp_admodds, join_by(name == term)) |>
  select(term = name, InvestigationDescription, odds, odds_adm, id, starts_with("tests"), p) |>
  mutate(InvestigationKey = as.numeric(str_extract(term, "[:digit:]{2}"))) |>
  left_join(
    lkp_invst_ec_aea |> 
      select(InvestigationKey, invst_group), 
    join_by(InvestigationKey)
  ) |> 
  # count(invst_group) |> 
  ggplot()+
  geom_hline(yintercept = 1, lty = "dashed", alpha = 0.1)+ # col = "grey40")+
  geom_vline(xintercept = 1, lty = "dashed", alpha = 0.1)+ # col = "grey40")+
  geom_point(aes(odds, odds_adm, size = p), alpha = 0.3, stroke = NA)+
  scale_size(labels = scales::percent_format())+
  geom_text_repel(
    aes(
      odds, odds_adm,
      label = str_wrap(InvestigationDescription, 20)
      ),
    size = 2,
  ) +
  facet_wrap(vars(invst_group))+
  theme_bw()+
  theme(
    axis.ticks = element_blank(),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7),
    axis.title = element_text(size = 7)
  )+
  labs(
    x = "\nCase-mix-adjusted odds of test (2023/24 vs 2019/20)",
    y = "\nCase-mix-adjusted odds of admission having received test (2023/24 vs 2019/20)\n",
    size = "Attendances\nwith test\n2023/24"
  )
```

# Discussion

[**N - SW**]{.underline}

# Possible Further Research

# Appendix 1 {#sec-appendix}

[**O - AJ**]{.underline}

Make all sections of the appendix collapsable?

Sampling and SQL queries ? Collapsable?

<!-- To carry out these analyses, we require only the EC data set, and only two years (provisionally, 2019/20 and 2023/24). We intend to use different samples to answer the different types of question). We will include a provider in a sample if the quality of their data submissions meet our requirements for the given question. We expect this approach to  -->

### Sampling for Q1 (from trends dq)

Given that we will be covering a relatively long period, spanning two data sets (with slightly different structures), it will be important to understand the quality and consistency of recording practices, both over time and by provider. In earlier [work in this area](https://www.strategyunitwm.nhs.uk/publications/waiting-times-and-attendance-durations-english-ae-departments), we based our analysis on subsets of providers who had a history of high quality data. For this analysis, it is likely that we will, again, have to work with data from a subset providers.

There is more - will have to look...

We identified XXX providers with provider codes that survived across the 12 year study period and that, in addition, passed basic data quality checks. These XXX providers represent the total population of providers that we might have used for our analysis had data quality been universally high.

Providers were included in our sample if their data submissions met the quality required for this question. Here, the study period is relatively long but we have few variables to examine. We are looking to see that:

1.  Throughout the study period, providers have had consistently high levels of recording of the investigation fields.

2.  For attendances in which multiple investigations of the same type have been carried out, providers have recorded each instance of the investigation (and have been consistent in this practice over the whole study period).

We established that XX providers met these criteria. The providers we intend to use for our sample are given in @tbl-sample-q1. In the adjacent tab, @tbl-representation suggests that this sample of XXX providers is widely representative of the population of providers. We note that the sample includes higher proportions of larger providers and providers in London, and lower proportions of smaller providers and providers in the East of England.

## Sampling for Q2 (from ECDS DATA qual qmd)

One solution to this problem may be to use other variables to understand casemix (and may even be more valid measures). For example, the combination of *acuity* (composed of 5 levels) and Chief Complaint Group (15 groups) should allow us to account for differences in casemix. Both variables have \<13% NULL values in both years when looking at the data set as a whole. The completion rates would likely improve further if we were to examine a subset of providers.

It is possible, then, that the quality of a provider’s submission of these two variables, along with *Der_EC_Investigation_All*, may determine whether we include the provider in our analyses. We will address this idea next.

we deduce that 16 providers can claim to have **all 3 fields** **complete** for \>90% of attendances, in both years. If we were to relax the threshold to \>85%, we get just four more providers.

### How representative is the provider sample?

We identified 102 providers with provider codes that were present in the two years of interest and that, in addition, passed basic data quality checks. These providers will represent the total population of providers that we might have used for our analyses. Given the data quality issues, we have selected a sample of providers to improve the reliability and validity of our results. However, since we wish to make general statements about about the effect of diagnostic testing on urgent and emergency care, we must ensure that our samples are representative of the population of providers.

More deprived, and older population, with more arrivals by ambulance results in more admissions. We will control for all of these factors - what difference does this make…?

# Appendix 2

::: {.callout-note collapse="true" appearance="minimal" icon="false"}
# Graphics (click to unhide)

**Tests per attendance**

```{r}
#| label: fig-tests-per-attd
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[1]] |>
  ggplot() +
  theme_bw() +
  # theme_minimal() +
  geom_vline(aes(xintercept = mean_tests, colour = fyear), linetype = "dashed") +
  geom_line(aes(x = n_tests_all, y = p_patients, colour = fyear)) +
  geom_point(aes(x = n_tests_all, y = p_patients, colour = fyear)) +
  scale_y_continuous(
    name = "Proportion of attendances",
    label = label_percent(accuracy = 1)
  ) +
  scale_x_continuous(name = "Number of tests") +
  labs(
    # title = "Number of tests per attendance",
    # subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24",
    caption = "Selected providers | Apr19-Feb20 & Apr23-Feb24\nDashed lines indicate mean number of tests"
  )
```

**Tests per attendance (grouped)**

```{r}
#| label: fig-grpd-tests-per-attd
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[2]] |>
  ggplot() +
  geom_col(
    aes(fyear, p_patients_adj, fill = n_tests_all_grp),
    position = position_stack()
  ) +
  geom_hline(aes(yintercept = 0), colour = "grey20", lty = "dashed") +
  geom_text(aes(x = fyear, y = p_patients_adj, group = n_tests_all_grp, label = p_patients_label),
    position = position_stack(vjust = 0.5)
  ) +
  # scale_fill_manual(values = c("red", "orange", "yellow", "grey")) +
  # scale_colour_manual(values = c("red", "orange", "yellow", "grey")) +
  scale_y_continuous(
    name = "Proportion of attendances",
    label = label_percent(accuracy = 1)
  ) +
  scale_x_discrete(name = "Financial year") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.y = element_blank()
  ) +
  labs(
    # title = "Number of tests per attendance",
    # subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24", 
    caption = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )
```

### Grouped by presentation and disposal

```{r}
#| label: fig-grpd-tests-per-attd-disp
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[3]] |>
  ggplot() +
  geom_col(aes(x = fyear, y = p_patients_adj, fill = n_tests_all_grp),
    position = position_stack()
  ) +
  geom_hline(aes(yintercept = 0), colour = "grey20", lty = "dashed") +
  geom_text(
    aes(x = fyear, y = p_patients_adj, group = n_tests_all_grp, label = p_patients_label),
    position = position_stack(vjust = 0.5),
    size = 2
  ) +
  facet_grid(cols = vars(is_admitted), rows = vars(inj_illness)) +
  scale_fill_manual(values = c("red", "orange", "yellow", "grey")) +
  scale_y_continuous(
    name = "Proportion of attendances",
    label = label_percent(accuracy = 1)
  ) +
  scale_x_discrete(name = "Financial year") +
  theme(legend.title = element_blank()) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Number of tests per patient by disposal and presentation type",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )
```

**Grouped by test type**

```{r}
#| label: fig-grpd-tests-per-attd-type
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[4]] |>
  ggplot() +
  geom_col(aes(x = fyear, y = p_patients_adj, fill = n_tests_grp),
    position = position_stack()
  ) +
  geom_hline(aes(yintercept = 0), colour = "grey20", lty = "dashed") +
  geom_text(aes(x = fyear, y = p_patients_adj, group = n_tests_grp, label = p_patients_label),
    position = position_stack(vjust = 0.5)
  ) +
  facet_wrap(vars(test_type)) +
  scale_fill_manual(values = c("orange", "yellow", "grey")) +
  scale_y_continuous(
    name = "Proportion of attendances",
    # label = label_percent(accuracy = 1),
    breaks = c(-0.5, -0.25, 0, 0.25, 0.5),
    labels = c("50%", "25%", "0%", "25%", "50%")
  ) +
  scale_x_discrete(name = "Financial year") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Number of tests per patient by test type",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )

```

### Time by presentation and disposal

```{r}


```

### Time by test type

```{r}
#| label: fig-time-vs-tests-grp-type
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| column: page
#| fig-width: 8
#| fig-asp: 0.618
#| out-width: "90%"

readRDS(here("data", "df_nested_preplots_multi_tests.rds"))$data[[7]] |>
  ggplot() +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  geom_line(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear
  )) +
  geom_point(aes(
    x = n_tests_trunc,
    y = median_dur_assess_depart,
    colour = fyear,
    size = n_patients
  )) +
  facet_wrap(vars(test_type)) +
  scale_y_continuous(
    name = "Median time in ED (assessment to departure)",
    limits = c(0, NA_real_)
  ) +
  scale_x_continuous(
    name = "Number of tests",
    limits = c(0, 5),
    breaks = (0:5),
    labels = c(
      "0", "1", "2", "3", "4",
      "5+"
    )
  ) +
  scale_size_continuous(name = "Number of attendances") +
  scale_colour_discrete(name = "Financial year") +
  labs(
    title = "Median time and number of tests in ED by test type",
    subtitle = "Selected providers | Apr19-Feb20 & Apr23-Feb24"
  )


```
:::
