---
title: "Data Quality in the 'Trends' Table"
subtitle: "Project: Diagnostic Growth UEC"
date: 2024-09-17
format:
  html:
    html-table-processing: none
    # code-fold: true
    code-tools: true
    embed-resources: true
    grid:
      margin-width: 350px
  pdf: default
reference-location: margin
citation-location: margin
toc: true
toc-title: Contents
toc-location: left
theme: journal
fig-cap-location: top
number-sections: true
title-block-banner: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library("gt")
library("gtExtras")
library("here")
library("dplyr")
library("tidyr")
library("purrr")
library("readr")
library("ggplot2")
library("janitor")
library("stringr")
library("lubridate")

# FOR GRAPHIC FYEAR AXIS LABELS:
every_nth <-  function(n) {
  return(function(x) {x[c(TRUE, rep(FALSE, n - 1))]})
}

# 1. LOAD DATASETS --------------------------------------------

# FOR REGION LKP:
eric_2223 <- read_csv("https://files.digital.nhs.uk/FB/BE3AC8/ERIC%20-%20202223%20-%20Trust%20data.csv") |> 
  clean_names() |> 
  mutate(region = str_remove_all(commissioning_region, " COMMISSIONING REGION")) |> 
  select(trust_code, trust_name, region, trust_type)

list.files(here::here("data_raw"))
dq_multicoding <- readRDS(
  here("data_raw", "from_ncdr_trends_240917_dq_multicoding.rds")
  )
ts_provider_attds <- readRDS(
  here("data_raw", "from_ncdr_240917_ts_prov_att.rds")
  )
dq_provider_na_24_99 <- readRDS(
  here("data_raw", "from_ncdr_trends_240917_dq_prov_na_24_99.rds")
  )
dq_provider_represent <- readRDS(
  here("data_raw", "from_ncdr_240917_prov_represent.rds")
  ) 
# DEPENDS ON RESULTS OF THIS DOCUMENT:
# df_preplot_trend <- readRDS(
#   here("data_raw", "from_ncdr_trends_240910_df_prep_plot.rds")
#   ) 

# 2. JOIN --------------------------------------------

dq_provider_represent <- dq_provider_represent |> 
  left_join(eric_2223, join_by(procode == trust_code))
```

# Background

In Phase I of the Diagnostic Growth project, we seek to answer five questions:

::: {.callout-note appearance="minimal"}
1.  How has the average number of investigations per ED attendance changed over time?

2.  For each of the 20+ test types, how much more likely is it that a patient will receive a test in ED in year 2023/24 than in year 2019/20?

3.  What proportion of the growth in each test, can be attributed to: (a) increases in raw demand; (b) changes in case-mix; and (c) changes in the propensity to test (i.e. clinical practice)?

4.  How much (net) time does a test add to a patient’s stay in ED?

5.  How do tests influence decisions to admit?
:::

In the first of these questions we will study a 12 year period, from 2012/13 to 2023/24. This will require the use of two different data sets: the AEA data set (to 2018/19), and the ECDS. The following table, found on the NCDR SQL server, gathers variables necessary for this analysis:

<!-- [NHSE_Sandbox_StrategyUnit].[dbo].[2232_diagnostics_trend] -->

```{sql connection = F, eval = F, error = T}
[NHSE_Sandbox_StrategyUnit].[dbo].[2232_diagnostics_trend]
```

The SQL code that created this table can be found in the [*sql* folder](https://github.com/The-Strategy-Unit/diagnostic_growth_uec/tree/phase1/sql/sql). Manipulation of the \[2232_diagnostics_trend\] table was undertaken on the NCDR DS server (see code in @sec-appendix) to produce the summary tables which serve as the bases for this analysis of data quality.

Given that we will be covering a relatively long period, spanning two data sets (with slightly different structures), it will be important to understand the quality and consistency of recording practices, both over time and by provider. In earlier [work in this area](https://www.strategyunitwm.nhs.uk/publications/waiting-times-and-attendance-durations-english-ae-departments), we based our analysis on subsets of providers who had a history of high quality data. For this analysis, it is likely that we will, again, have to work with data from a subset providers.

# Ensuring data quality {#sec-two}

In the analysis of trends, we have relatively few variables to examine. And perhaps our sole interest regarding data quality in this case is the *Der_Investigation_All* field (from AEA) and the *Der_AEA_Investigation_All* field (from ECDS). Specifically we are looking to see that:

1.  Throughout the study period, providers have had consistently high levels of recording of the field, *Der\_(AEA)\_Investigation_All*.

2.  For attendances where multiple investigations of the same type have been carried out, providers have recorded each instance of the investigation (and have been consistent in this practice over the whole study period).

    While there is no way to tell if the recorded number of tests matches the actual number of tests, we can, at least, identify providers whose data suggests that no patient received more than one of a given test. (It is highly improbable, for example, that over the course of one year, not one patient attending an ED would be given more than one biochemistry test. This, instead, indicates that a provider's practice is to record one biochemistry test, no matter how many of these tests were carried out.)

    It's possible that this point may be of more interest for some tests (e.g. CTs) than others (e.g. biochemistry tests). What is the difference in time demands of multiple biochemistry tests versus multiple CTs?

    We also see two growth-related questions here:

    i\) Are more tests being carried out? And,

    ii\) Has the distribution of tests changed?

    We will plan to address the former, for now, though the latter question may also be of interest.

The analysis that follows will explore issues raised in the two points above. Data manipulation was conducted within R on the NCDR Data Science Server (see Appendix for code), and summary data frames were extracted. The inquiries stem from these summary data frames.

## Selecting a study population on the basis of provider data quality

```{r}
#| label: provider-survival
#| include: false

ts_provider_survival <-  ts_provider_attds |>
  count(fyear) 

providers_surviving <- ts_provider_attds |>
  # MORE THAN AV OF 15 A DAY FOR A YEAR (EXCLUDES RVR ANOMALY)
  filter(n > 15*365) |>
  # NOW CLOSED (FROM MID 2023):
  filter(procode != "RVY") |>
  complete(fyear, procode) |>
  group_by(procode) |>
  filter(!is.na(sum(n))) |>
  ungroup() |>
  distinct(procode)

ts_provider_attds |> 
  semi_join(providers_surviving, join_by(procode)) |> 
  arrange(procode) |> 
  # slice(1:(33*12)) |> 
  # slice(((33*12)+1):(66*12)) |> 
  slice(((66*12)+1):(99*12)) |> 
  ggplot() +
  geom_blank(aes(y = 0)) +
  geom_line(aes(fyear, n, group = procode)) +
  scale_x_discrete(breaks = every_nth(n=7))+
  facet_wrap(vars(procode), scales = "free_y")
  
```

We identified `r nrow(providers_surviving)` providers with provider codes that survived across the 12 year study period and that, in addition, passed basic data quality checks. These `r nrow(providers_surviving)` providers will represent the total population of providers that we might have used for our analysis had data quality been universally high.

What follows is a sensitivity analysis to examine how different interpretations of "multiple tests" - and different thresholds for the frequency of these cases - might affect the number of providers we use in our analysis of trends.

We will start with data quality point 2, in @sec-two above, and seek to understand how common it has been for providers to report multiple tests of the same type in a single record. We will look at 6 test types (see footnote 1), and at cases in which there are two, three, and four instances of each of these tests within a record. We will see how many providers record multiple investigations, and how frequently these cases occur.

```{r}
#| label: mcs
#| include: false

dq_multicoding_sensitivity <- dq_multicoding |>
  cross_join(
    tibble(n_tests_of_type_in_record = c(2, 3, 4))
  ) |>
  cross_join(
    # MINIMUM PROPORTION OF CASES (AS PROPORTION OF 1-4) IN EACH YEAR:
    tibble(annual_instance_threshold = c(
      1 / 100000,
      1 / 50000,
      1 / 10000,
      1 / 5000,
      1 / 1000,
      1 / 500,
      1 / 100,
      1 / 50,
      1 / 10
    ))
  ) |>
  mutate(data = pmap(
    list(data, n_tests_of_type_in_record, annual_instance_threshold),
    \(df, y, z)
    df |>
      mutate(procode = str_sub(procode, 1, 3)) |> 
      filter(code_occurs == y) |>
      # WE WILL GET PRACTICE VARIATION ACROSS PROVIDERS, SO WHAT P DO WE PICK?
      # (% CASES IN ALL YEARS > z)
      filter(p > z) |>
      # NEXT FEW LINES ARE TO MAKE SURE ALL YEARS ACCOUNTED FOR:
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup()
  ))

```

```{r}
#| label: mcs-provider-n
#| include: false

dq_multicoding_provider_n <- dq_multicoding_sensitivity |>
  mutate(n_prov = map_vec(
    data,
    \(df)
    df |>
      distinct(procode) |>
      count() |>
      pull(n)
  )) |>
  mutate(n_prov_500 = map_vec(
    data,
    \(df)
    # MORE THAN 500 CASES EACH YEAR:
    df |>
      filter(n >= 500) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |>
  mutate(n_prov_50 = map_vec(
    data,
    \(df)
    # MORE THAN 50 CASES EACH YEAR:
    df |>
      filter(n >= 50) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |> 
  mutate(n_prov_20 = map_vec(
    data,
    \(df)
    # MORE THAN 20 CASES EACH YEAR:
    df |>
      filter(n >= 20) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |> 
  mutate(n_prov_2 = map_vec(
    data,
    \(df)
    # MORE THAN 1 CASE EACH YEAR:
    df |>
      filter(n >= 2) |>
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |>
      count(procode) |>
      count() |>
      pull(n)
  )) |> 
  mutate(n_tests_of_type_in_record = as.character(n_tests_of_type_in_record)) |>
  mutate(invest_name = case_when(
    invest_code == "01" ~ "01 | X-ray",
    invest_code == "03" ~ "03 | Haematology",
    invest_code == "05" ~ "05 | Biochemistry",
    invest_code == "10" ~ "10 | Ultrasound",
    invest_code == "11" ~ "11 | MRI",
    invest_code == "12" ~ "12 | CT"
  ))
```

We might use @fig-mcs to identify providers who have investigation recording practices that meet our needs. @fig-mcs can be interpreted as outlined in the following example: The top left panel concerns X-rays. The red line represents cases in which two X-rays have been recorded in an attendance. As we increase the proportion of attendances with 2 X-rays (as a share of *all attendances with an X-ray*), by moving along the horizontal axis (from 1 in 100,000, to 1 in 10), the number of providers meeting the threshold falls from over 40 to 20. The graphic should provide some indication of how rarely an event occurs across providers.

::: callout-note
The horizontal axis in @fig-mcs is scaled logarithmically.
:::

```{r}
#| label: fig-mcs
#| echo: false
#| fig-cap: "Coding of multiple investigations: A sensitivity analysis"

dq_multicoding_provider_n |> 
  ggplot()+
  theme_bw()+
  geom_line(aes(annual_instance_threshold, n_prov, col = n_tests_of_type_in_record, group = n_tests_of_type_in_record))+
  geom_point(aes(annual_instance_threshold, n_prov, col = n_tests_of_type_in_record, group = n_tests_of_type_in_record))+
  facet_wrap(vars(invest_name))+
  # scale_x_continuous(labels = c("1/1000", "1/100", "1/50", "1/33", "1/25", "1/20"))+
  scale_x_log10(
    breaks = c(
    1/100000,
    1/10000,
    1/1000,
    1/100,
    1/10
  ),
    labels = c(
    "1/100000",
    "1/10000",
    "1/1000",
    "1/100",
    "1/10"
  )
  )+
  theme(
    legend.position = "top",
    axis.text = element_text(size = 5),
    axis.title = element_text(size = 8),
    axis.ticks = element_blank()
    )+
  labs(
    x = "constitutes at least x records (of all of records with > 0 instances of that investigation type) in each and every year",
    y = "number of providers")

# 2 X-rays constitute (1/1000) records for x providers
# 2 X-rays constitute (1/500) records for y providers

```

Additionally, @fig-mcs shows that patients were far more likely to receive multiple tests for some modalities than they were for others. Multiple biochemistry tests (2, 3, and 4 tests per attendance) were common throughout the study period. Conversely, few providers gave a patient multiple MRIs throughout the period (and, even then, these cases were uncommon). The same appears to be true of ultrasounds. We will therefore exclude these two from the rest of this sensitivity analysis.

::: column-margin
It's worth noting, moreover, that the availability of some of these tests has grown rapidly over the period. This is especially true of newer technologies like CT and MRI. For these modalities in particular, we may wish to be more liberal about the frequency of cases we accept, since - for example - multiple MRIs (and MR scanners themselves) would have been far less common in 2012 than in 2023. Our thresholds represent the minimum number of cases in each year. Thus, given the growing prevalence of these technologies, the bar would be set high early in the period (relative to the level late in the period).
:::

### Option 1: Strict interpretation of coding practices

If we wish to take a strict line about coding practices relating to multiple tests of the same type, then we might assume that providers who code **three** tests of the same type would always record multiple tests of the same type. (We will choose three since it represents a higher bar, and because we would expect that codes recorded in triplicate would less likely be errors than codes duplicated.)

We can then select a (relatively liberal) case proportion threshold for each of the tests. (A reminder that the threshold (in this case) is the share of *all attendances with a given test type* composed of 3 tests of that type. The threshold gives and indication of how rarely such an event occurs.)

| Investigation Code | Investigation Type | Case Proportion Threshold |
|--------------------|--------------------|---------------------------|
| 01                 | X-ray              | 1/5000                    |
| 03                 | Haematology        | 1/5000                    |
| 05                 | Biochemistry       | 1/500                     |
| 12                 | CT                 | 1/5000                    |

```{r}
#| include: false

dq_multicoding_provider_lists <- dq_multicoding_provider_n |> 
  filter(
    (invest_code == "01" & n_tests_of_type_in_record == 3 & annual_instance_threshold == 1/5000)|
    (invest_code == "03" & n_tests_of_type_in_record == 3 & annual_instance_threshold == 1/5000)|
    (invest_code == "05" & n_tests_of_type_in_record == 3 & annual_instance_threshold == 1/500)|
    (invest_code == "12" & n_tests_of_type_in_record == 3 & annual_instance_threshold == 1/5000)
  ) |> 
  mutate(prov_list = map(
    data,
    \(df)
    df |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
  mutate(prov_list_500 = map(
    data,
    \(df)
    df |> 
      filter(n >= 500) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
  mutate(prov_list_10 = map(
    data,
    \(df)
    df |> 
      filter(n >= 10) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    ))

dq_providers_conservative <-
  
  # AT LEAST 20 RECORDS EACH AND EVERY YEAR:
  semi_join(
    providers_surviving,
  reduce(dq_multicoding_provider_lists$prov_list_10, semi_join),
  join_by(procode)
  )
  
# ,
  # AT LEAST 500 RECORDS EACH AND EVERY YEAR:
#   dq_multicoding_provider_lists$prov_list_500[[3]],
#   join_by(procode)
# )
# dq_providers_conservative <-
#   semi_join(
#   # AT LEAST 20 RECORDS EACH AND EVERY YEAR:
#   reduce(dq_multicoding_provider_lists$prov_list_20[c(1,2,4)], semi_join),
#   # AT LEAST 500 RECORDS EACH AND EVERY YEAR:
#   dq_multicoding_provider_lists$prov_list_500[[3]],
#   join_by(procode)
# )
```

In this strict approach, we will only include providers in our analysis of trends if:

-   They meet the threshold level (for coding of three identical tests) across all four test types.

-   The threshold level equates to **at least 10 attendances**. (Note: setting this too high will discriminate against smaller providers.)

This first step would leave us with `r nrow(dq_providers_conservative)` providers:

`r dq_providers_conservative$procode`.

We will now assess data quality point 1 (from @sec-two), and use @fig-dq-ts-strict to check how well these providers have recorded investigation codes over time. Analysis suggests that around 70% of patient attendances will involve a diagnostic test, and that this value has not changed appreciably over time. We will therefore use this value as a reference level, which will serve to highlight anomalies.

```{r}
#| label: fig-dq-ts-strict
#| echo: false
#| message: false
#| fig-cap: "Timeseries Showing Test Coding Practices for Providers in Sample Option 1"

dq_provider_na_24_99 |>
  semi_join(dq_providers_conservative, join_by(procode)) |>
  ggplot() +
  geom_area(aes(fyear, p, fill = code, group = code), alpha = 0.6, col = "white") +
  geom_hline(yintercept = 0.7, lty = "dashed", alpha = 0.6) +
  scale_x_discrete(breaks = every_nth(n=7))+
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    legend.position = "top"
  ) +
    labs(y = "% of records with code")+
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(vars(procode))+
        guides(fill = guide_legend(reverse = TRUE))
```

### Option 2: Relaxed interpretation of coding practices

A relaxed interpretation of coding practices relating to multiple tests of the same type should increase the population covered by the study. Here, we might assume that providers who record **two** tests of the same type in an attendance record would always record multiple tests of the same type. To offset this (perhaps slightly weaker) assumption, we might set conservative threshold for each of the tests. (A reminder that the threshold (in this case) is the share of *all attendances with a given test type* composed of 2 tests of that type.)

| Investigation Code | Investigation Type | Case Proportion Threshold |
|--------------------|--------------------|---------------------------|
| 01                 | X-ray              | 1/1000                    |
| 03                 | Haematology        | 1/500                     |
| 05                 | Biochemistry       | 1/100                     |
| 12                 | CT                 | 1/1000                    |

In this relaxed approach, we will include providers in our analysis of trends if:

-   They meet the threshold level (for coding of two identical tests) across all four test types.

-   The threshold level equates to **at least 5 attendances**. (Note: setting this too high will discriminate against smaller providers.)

```{r}
#| include: false

dq_multicoding_provider_lists_relaxed <- dq_multicoding_provider_n |> 
  filter(
    (invest_code == "01" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/1000)|
    (invest_code == "03" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/500)|
    (invest_code == "05" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/100)|
    (invest_code == "12" & n_tests_of_type_in_record == 2 & annual_instance_threshold == 1/1000)
  ) |> 
  mutate(prov_list = map(
    data,
    \(df)
    df |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
  mutate(prov_list_500 = map(
    data,
    \(df)
    df |> 
      filter(n >= 500) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
  mutate(prov_list_50 = map(
    data,
    \(df)
    df |> 
      filter(n >= 50) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
   mutate(prov_list_20 = map(
    data,
    \(df)
    df |> 
      filter(n >= 20) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    )) |> 
   mutate(prov_list_4 = map(
    data,
    \(df)
    df |> 
      filter(n >= 5) |> 
      complete(fyear, procode) |>
      group_by(procode) |>
      filter(!is.na(sum(code_occurs))) |>
      ungroup() |> 
      distinct(procode) 
    ))

# dq_providers_relaxed <-
#   semi_join(
#   # AT LEAST 500 RECORDS EACH AND EVERY YEAR:
#   reduce(dq_multicoding_provider_lists_relaxed$prov_list_500[1:3], semi_join),
#   # AT LEAST 50 RECORDS EACH AND EVERY YEAR:
#   dq_multicoding_provider_lists_relaxed$prov_list_50[[4]],
#   join_by(procode)
# )

dq_providers_relaxed <-
  # semi_join(
  # AT LEAST 2 RECORDS EACH AND EVERY YEAR:
  semi_join(
    providers_surviving,
  reduce(dq_multicoding_provider_lists_relaxed$prov_list_4, semi_join),
  join_by(procode)
  )
  
  # ,
#   # AT LEAST 50 RECORDS EACH AND EVERY YEAR:
#   dq_multicoding_provider_lists_relaxed$prov_list_50[[4]],
#   join_by(procode)
# )

```

This first step would leave us with `r nrow(dq_providers_relaxed)` providers:

`r dq_providers_relaxed$procode`.

We will now assess data quality point 1 (from @sec-two), and use @fig-dq-ts-relax to check how well these providers have recorded investigation codes over time. Analysis suggests that around 70% of patient attendances will involve a diagnostic test, and this figure has not changed appreciably over time. We will therefore use this figure as a reference level, which will serve to highlight anomalies.

```{r}
#| label: fig-dq-ts-relax
#| echo: false
#| message: false
#| fig-cap: "Timeseries Showing Test Coding Practices for Providers in Sample Option 2"

dq_provider_na_24_99 |>
  semi_join(dq_providers_relaxed, join_by(procode)) |>
  ggplot() +
  geom_area(aes(fyear, p, fill = code, group = code), alpha = 0.6, col = "white") +
  geom_hline(yintercept = 0.7, lty = "dashed", alpha = 0.6) +
  scale_x_discrete(breaks = every_nth(n=7))+
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    legend.position = "top"
  ) +
  labs(y = "% of records with code")+
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(vars(procode))+
        guides(fill = guide_legend(reverse = TRUE))
```

## How representative is provider sample option 2?

We identified `r nrow(providers_surviving)` providers with provider codes that survived across the 12 year study period and that, in addition, passed basic data quality checks. These `r nrow(providers_surviving)` providers will represent the total population of providers that we might have used for our analysis.

Given our data quality requirements, we will base our trend analysis on the sample of `r nrow(dq_providers_relaxed)` providers identified in option 2. We expect this approach will improve the reliability and validity of our results. Yet, since we also wish make general statements about changes in test rates, we must check that our sample is representative of the population of providers. @tbl-representation suggests that the sample of `r nrow(dq_providers_relaxed)` providers, identified above, is widely representative of the population of providers. We note that the sample includes higher proportions of larger providers and providers in London, and lower proportions of smaller providers and providers in the East of England.

<!-- With the trends query where clause and the addtional fields that we need for verifying representation New query, same where clause (with all ECDS years) fields: We will take the relaxed sample For 2023/24 (or average over period?) -->

```{r}
#| label: tbl-representation
#| tbl-cap: "Is the Sample Representative of the Wider Provider Population?"
#| echo: false
#| message: false

# ARE WE COMPARING WITH xx WHO LASTED THROUGHT STUDY PERIOD

table_representation <- tibble(
  data = list(
    dq_provider_represent |>
      # semi_join(low_attendance_filter, join_by(procode)) |>
      semi_join(providers_surviving, join_by(procode)),
    dq_provider_represent |>
      semi_join(dq_providers_relaxed, join_by(procode))
  )
) |>
  mutate(n_providers = map_dbl(data, \(df)
  df |>
    count(procode) |>
    nrow())) |>
  # THIS WILL BE SIZE IN 2023/24:
  # (FOR OTHER VARS WE'LL LOOK AT MEAN ACROSS 5 YR ECDS TIME PERIOD)
  mutate(size = map(data, \(df)
  df |>
    filter(fyear == "2023/24") |> 
    count(fyear, procode, wt = n) |>
    group_by(procode) |>
    # MEAN OF A PROVIDER'S SIZE OVER 5 YEARS:
    reframe(mean_n = mean(n)) |>
    mutate(size_group = case_when(
      mean_n < 70e3 ~ "att < 70k",
      mean_n >= 70e3 & mean_n < 95e3 ~ "att 70-94k",
      mean_n >= 95e3 & mean_n < 130e3 ~ "att 95-129k",
      mean_n >= 130e3 ~ "att > 130k",
      T ~ NA_character_
    )) |>
    mutate(size_group = factor(
      size_group,
      levels = c("att < 70k", "att 70-94k", "att 95-129k", "att > 130k")
    )) |>
    count(size_group) |>
    mutate(p = n / sum(n)) |>
    select(-n) |>
    pivot_wider(names_from = size_group, values_from = p)
  
  )) |>
  #
  # mutate(min_size = map_dbl(data, \(df)
  #                          df |>
  #                            count(fyear, procode, wt = n) |>
  #                            summarise(av = round(min(n))) |>
  #                            pull(av)
  # )) |>
  #
  # mutate(max_size = map_dbl(data, \(df)
  #                          df |>
  #                            count(fyear, procode, wt = n) |>
  #                            summarise(av = round(max(n))) |>
  #                            pull(av)
  # )) |>
  # mutate(av_size = map_dbl(data, \(df)
  #                          df |>
  #                            count(fyear, procode, wt = n) |>
  #                            summarise(av = round(mean(n))) |>
  #                            pull(av)
  # )) |>
  # mutate(type = map(data, \(df)
  #                   df |>
  #                     count(fyear, trust_type, wt = n, sort = T) |>
  #                     group_by(fyear) |>
  #                     mutate(p = n/sum(n)) |>
  #                     group_by(trust_type) |>
  #                     summarise(av = round(mean(p), 3)) |>
  #                     ungroup() |>
  #                     pivot_wider(names_from = trust_type, values_from = av) |>
  #                     clean_names() |>
  #                     rename_with(~ paste0("type_", .x, recycle0 = TRUE)) |>
  #                     identity()
  # )) |>
  mutate(arrive_ambulance = map_dbl(data, \(df)
  df |>
    count(fyear, is_ambulance, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_ambulance == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(admitted = map_dbl(data, \(df)
  df |>
    count(fyear, is_adm, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_adm == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(age_under_18 = map_dbl(data, \(df)
  df |>
    count(fyear, age_under_18, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(age_under_18 == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(age_75_plus = map_dbl(data, \(df)
  df |>
    count(fyear, age_75_plus, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(age_75_plus == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(urban = map_dbl(data, \(df)
  df |>
    count(fyear, is_urban, wt = n) |>
    # BECAUSE LOTS OF NAS IN 2023/24:
    filter(fyear != "2023/24") |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(is_urban == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(deprived_20pc = map_dbl(data, \(df)
  df |>
    count(fyear, imd_quint_1, wt = n) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    ungroup() |>
    filter(imd_quint_1 == 1) |>
    summarise(av = round(mean(p), 3)) |>
    pull(av))) |>
  mutate(region = map(data, \(df)
  df |>
    count(fyear, region, wt = n, sort = T) |>
    group_by(fyear) |>
    mutate(p = n / sum(n)) |>
    group_by(region) |>
    summarise(av = round(mean(p), 3)) |>
    ungroup() |>
    pivot_wider(names_from = region, values_from = av) |>
    clean_names() |>
    rename_with(~ paste0("region_", .x, recycle0 = TRUE)) |>
    identity())) |>
  unnest(region) |>
  unnest(size) |>
  # unnest(type) |>
  select(-data) |>
  mutate(zz = if_else(n_providers > 50, "population", "sample"), .before = n_providers) |>
  pivot_longer(cols = !starts_with("zz"), names_to = "variable") |>
  pivot_wider(names_from = zz, values_from = value)

table_representation |> 
  select(variable, population, sample) |> 
  mutate(across(2:3, ~if_else(is.na(.), 0, .))) |> 
  rowwise() |> 
  mutate(diff = sample - population) |> 
  gt() |> 
  opt_row_striping(row_striping = F) |> 
  fmt_auto() |> 
  tab_style(
    style = list(
      cell_text(transform = "capitalize", weight = "bold")
    ),
    # different location
    locations = cells_column_labels(everything())
  ) |> 
  fmt_percent(rows = c(2:18), decimals = 1) |>
  tab_row_group(
    label = "Patient demographics:",
    rows =  str_detect(variable, "age|urban|depriv|region_")
  ) |> 
  tab_row_group(
    label = "Case-mix-related:",
    rows =  str_detect(variable, "arrive_|admitted")
  ) |> 
  tab_row_group(
    label = "Provider Size:",
    rows =  str_detect(variable, "att")
  ) |> 
  # tab_row_group(
  #   group = "Provider-related",
  #   rows =  str_detect(variable, "type_|min_|max_|av_")
  # ) |> 
  tab_row_group(
    label = "",
    rows =  str_detect(variable, "n_prov")
  ) |>
  tab_style(
  style = cell_text(weight = "bold"),
  locations = cells_row_groups(groups = c(1, 2, 3, 4))
) |> 
    sub_values(values = -71, replacement = "") |> 
   gt_highlight_cols(sample, fill = "grey", alpha = 0.1) |> 
  tab_options(data_row.padding = px(1))

```

<!-- Previous sample was not representative of smaller sized providers. ///Who are the smaller providers??? What/where are they ....childrens..? Why does multicoding exclude them particularly? ///Smaller providers and multicoding - we'd always expect them to have had biochem. /// but CT as a newer tech - perhaps not in earlier years??? Perhaps no CT at all for them at that stage. /// Do smaller providers have coding quality problems? This seems possible.  /// OR smaller providers -  - less serious cases - multiple CTs less common adjust thresholds /// OF COURSE for smalller providers, nartural variation more influential on criteria.... AND the requirement of a certain number ALSO excludes smaller providers. /// BASEe the N on the size of provider... or just remove the n condition because you have the proportion. /// WHAT sort of SIZE are they - when we are talking of small????? ///  START WITH SIMPLEST - RULE OUT - MOVE TO MORE COMPLEX /// TODO: start here and rule these out /// Quick examine of smaller providers (what type, size, over years) /// Work out what removing the n criteria means in terms of size... /// and remove n equivalent for threshould criterion /// Smaller providers deep dive. (incl. are CTs seen  in earlier years) /// look at later years for CT /// take only biochemisty --->

# Estimated trend in ED test rate

<!--  Based on the ... providers identified in sample option 2. --->

```{r}
#| include: false
##| label: fig-trend
##| echo: false
##| fig-cap: "Estimated average number of tests per ED attendance from #2012/13 to 2023/24, based on the 28 providers identified in sample option #2"
##| message: false



# you want to put the selection in here. 

# df_preplot_trend |> 
#   mutate(disdest = case_when(
#     disdest == "overall (all discharges)" ~ "overall (all destinations)",
#     disdest == "admitted" ~ "admitted patients",
#     disdest == "discharged" ~ "discharged patients",
#     T ~ disdest)) |> 
#  #  mutate(disdest = if_else(disdest == "overall (all discharges)", "overall (all destinations)", disdest)) |> 
#   # mutate(fyear_char = fyear) |> 
#   # mutate(fyear = as.numeric(str_sub(fyear, 6,7))) %>%
#   ggplot(aes(fyear, rate)) +
#   # geom_smooth(
#   #   # data = . %>% filter(fyear != 21),
#   #   aes(fyear, rate), method = "lm",
#   #   # se = F,
#   #   col = "grey80", lty = "dotted")+ #
#   geom_line(aes(col = disdest, group = disdest)) +
#   geom_point(aes(col = disdest, group = disdest)) +
#   geom_blank(aes(y = 0)) +
#   scale_color_manual(values = c("indianred3", "dodgerblue", "purple4")) +
#   labs(x = "Financial year")+
#   theme_bw()+
#   labs(y = "Number of tests per recorded ED attendance")+
#   theme(
#     legend.position = "none",
#     # axis.title.y = element_blank()
#     axis.ticks = element_blank()
#     )+
#   facet_wrap(vars(disdest))+ # , scales = "free_y"
#   # scale_x_continuous()+
#   scale_x_discrete(breaks = every_nth(n=4))+
#   scale_y_continuous(breaks = seq(0,9, by = 2))
#   
# #   
# # df_trend_by_invest |> saveRDS("from_ncdr_trends_240909_df_by_invst_prep_plot.rds")
# # 
# # tmp1 |> 
# #   group_by(fyear, disdest, invst_type) |> 
# #   reframe(n_att = sum(n_att, na.rm = T), n_invst = sum(n_invst, na.rm = T)) |> 
# #   # count(disdest)
# #   # filter(disdest == "discharged") |> 
# #   filter(disdest == "admitted") |> 
# #   mutate(rate = n_invst/n_att) |> 
# #   ggplot(aes(fyear, rate)) +
# #   geom_line(aes(col = disdest, group = disdest)) +
# #   geom_point(aes(col = disdest, group = disdest)) +
# #   geom_blank(aes(y = 0)) +
# #   # scale_color_manual(values = c("grey80", "grey50", "grey20")) +
# #   labs(x = "Financial year")+
# #   theme_bw()+
# #   # labs(subtitle = "Vertical axis: Average number of investigations per A&E attendance")+
# #   theme(legend.position = "none", axis.title.y = element_blank())+
# #   facet_wrap(vars(invst_type), scales = "free_y")+ # , scales = "free_y"
# #   scale_x_discrete(breaks = every_nth(n=3))+
# #   # scale_y_continuous(breaks = seq(0,9, by = 2))+
# #   NULL
# # 
# #   
# # 
# # 

```

# Appendix {#sec-appendix}

R code run on NCDR Data Science Server:

```{r}
#| label: appendix
#| include: false
#| eval: false
#| error: true

# SEE SCRIPT dq_trends.R
```
